---
title: "Lab 05 (alternate)"
#author: "Dr. Patrick Bitterman"
#date: "`r format(Sys.time(), '%Y %B %d')`"
output: pdf_document
my_subtitle: "Env Analysis in R"
fancy: true
geometry: margin=1in
latex_engine: pdflatex
colorlinks: true
---

# Lab 05: Making static maps

### Read the instructions COMPLETELY before starting the lab

This lab builds directly on the in-class frankenmap exercise and your readings from Lovelace, Ch. 9 (https://r.geocompx.org/adv-map.html). You should have the in-class notebook available for reference.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

## Introduction

In class we built the "frankenmap" as a group exercise — three teams working on pieces that were assembled into a single visualization. This lab asks you to build that same map on your own, but in a more deliberate way: **piece by piece, verifying each component before moving to the next**.

The key insight from the in-class exercise is that `tmap` is designed around composable pieces. Every map is just a base shape plus layers plus aesthetics:

```
map <- tm_shape(data) + tm_polygons() +   # base
       tm_shape(data2) + tm_lines() +     # next layer
       tm_compass() + tm_scalebar()       # layout elements
```

And maps stored as variables can be added together:

```
map_full <- map_part_1 + map_part_2 + map_part_3
```

This lab walks through each component of the frankenmap systematically. By the end of Task 4, you will have a complete inset map that integrates all of them. Task 5 asks you to produce a second, entirely original map.

All data are in `../data/static_mapping/` relative to this file's location.

```{r packages, echo=TRUE, message=TRUE}
library(sf)
library(terra)
library(tmap)
library(tidyverse)
library(grid)  # needed for inset maps in Task 4 — part of base R, no install required
library(stringr)
```

---

## Task 1: The state scale — Ohio choropleth

*Reference: Lovelace Ch 9.2.3–9.2.4 (color settings, choropleth mapping)*

In the frankenmap, Group 1 was responsible for a state-scale choropleth of Ohio counties. You will build that component here.

### 1.1 — Load and join

Load the Ohio county geometry from `oh_counties.gpkg`. Also load the 2020 Census demographic data from `oh_counties_DP2020.csv`.

```{r load_counties, echo=TRUE, message=TRUE}
#ohio counties geometry
allohio <- sf::read_sf("./static_mapping/oh_counties.gpkg") #geopackage

#2020 Census data
census <- read_csv("./static_mapping/oh_counties_DP2020.csv")
```

Before joining, inspect the GEOID column in **both** objects. You should notice something about the format of the GEOID in the CSV that requires attention before a join will work. Resolve this discrepancy, then perform a `left_join()`.

*Hint: Try `head()` on both GEOIDs and compare carefully.*

```{r join_data, echo=TRUE, message=TRUE}
#check out geoid
head(allohio)
head(census) #just need the last 5 digits from GEOID

census <- census %>% 
  mutate(GEOID = stringr::str_sub(geoid, 10, 15)) #extracted the right digits for state and county

oh_census <- left_join(allohio, census, by="GEOID")
```

Verify the join worked: how many rows does your joined object have? Does it match what you'd expect?
-This join is what I suspect because the census data had one entry for just the state of ohio, which is not included in the allohio geopackage that I loaded in. This join accounts for all counties being merged together (88)


### 1.2 — Choose and justify a variable

Choose **one** demographic variable from the Census data. If it is a raw count, normalize it appropriately (e.g., as a proportion of total population). Add a brief comment in your code explaining what the variable represents and why you find it interesting.

```{r choose_variable, echo=TRUE, message=TRUE}
#I will look at female population, normalized by the entire population of males and females per county
# I find the female population ratio interesting because I have lived in Kent for about 4 years now and there is a large female population here because of Kent State's programs and I am curious what other counties may have a high female population (perhaps to live there in the future as well).
oh_census <- oh_census %>%
  mutate(f_prop = (female/poptotal)) #proportion of f:totalpop
```

### 1.3 — Build the Ohio-scale map

Using your joined data, build a choropleth map of all Ohio counties. Requirements:

- Color the counties by your chosen variable using an **explicit** classification scheme — do not leave it at the default. Choose from `"equal"`, `"jenks"`, `"quantile"`, or `"pretty"` and set your number of classes.
- Style the borders with non-default line properties (adjust at least one of: color, width, or line type)
- Add a scale bar
- Give your legend a meaningful, human-readable title (not just the raw variable name)
- Store the finished map in a variable called `map_ohio`

```{r map_ohio, echo=TRUE, message=TRUE}
#choropleth map!!
map_ohio <- tm_shape(oh_census) + tm_polygons(fill="f_prop", style="jenks", palette="brewer.purples",
                                  title = "Female Population Proportion") + #title and pretty palette
  tm_borders(col="darkgrey", lwd=2) + #border color and line width change
  tm_compass(type = "8star", position = c("right", "bottom")) +
  tm_scalebar(position = c("right", "bottom")) 
  

# verify it looks right:
map_ohio
```

**Reflection question 1 (answer in your repo):** Justify your choice of classification method. How would the map's message change if you used a different method? Reference specific class boundaries in your answer.
-I looked at all of the style versions and decided to go with Jenks because it shows different groups of similar values in the data and maximizes the differences between those groups. Since I am looking at population, I wanted to see larger differences between that breaks of population proportion. Also the jenks style included the break at .5, which really shows what areas are greater than or equal to half the total population. The quantile style was my second choice, but it didn't clear distinguish when counties were meeting that .5 mark or higher.
---

## Task 2: The local scale — Portage and Summit counties

*Reference: Lovelace Ch 9.2.2–9.2.3 (map objects, aesthetics)*

Group 2 was responsible for the detail view: municipal boundaries, parks, and water features for the Portage and Summit county study area. You will build this component here, layer by layer.

### 2.1 — Prepare the study area

Filter the county geometry to Portage and Summit counties. Store the result as `sa`. You will also need a bounding box polygon of the study area for use in the inset map later — create it using `sf::st_bbox()` and `sf::st_as_sfc()` and store it as `oh_region`.

```{r study_area, echo=TRUE, message=TRUE}
#creating study area
sa <- allohio %>% dplyr::filter(NAME %in% c("Portage", "Summit")) %>% st_make_valid()

#turn this into a bounding box
oh_region <- sf::st_as_sfc(st_bbox(sa))

#check bbox
plot(oh_region)
```

### 2.2 — Municipal boundaries with labels

Load `oh_places.gpkg`. Before doing any spatial operations, check the CRS against `sa` and reproject if necessary. Spatially filter to the study area using `sf::st_intersection()`.

Build a map layer showing municipal boundaries with text labels for place names. In `tmap`, labels can be added using `tm_text()` as a separate layer, or via the `labels` argument in some geometry functions.

```{r places, echo=TRUE, message=TRUE}
# load in oh_places
oh_places <- sf::read_sf("./static_mapping/oh_places.gpkg") %>%
  st_make_valid() #fix the validity issue later on

#check crs
st_crs(oh_places) == st_crs(sa) #yippee they match

#filtering to just the study area
places_sa <- sf::st_intersection(sa, oh_places)

#tmap
places_map_nolab <- tm_shape(places_sa) + 
              tm_polygons(fill="#F8EDBE") +
              tm_borders(col="black", lwd=1.5) 
              #tm_labels_highlighted(text="NAME.1")
              #tm_text("NAME.1", 
                     # options = opt_tm_text(shadow=TRUE, remove_overlap = TRUE),
                     # size = 0.9,  fontface = "bold")
places_map <- tm_shape(places_sa) + 
              tm_polygons(fill="lightgrey") +
              tm_borders(col="black", lwd=1.5) +
              tm_text("NAME.1", 
                     options = opt_tm_text(shadow=TRUE, remove_overlap = TRUE),
                     size = 0.9,  fontface = "bold")
places_map
```

*Note: You may want to adjust label size or use `tm_text(size = ...)` to avoid overlap. This often requires experimentation.*

### 2.3 — Parks symbolized by type

Load `oh_parks.gpkg`. Reproject to match `sa` and spatially intersect. Symbolize the parks by their `FEATTYPE` attribute using **shades of green**. You have discretion over the exact palette — consult the `tmap` documentation or ColorBrewer for options.

```{r parks, echo=TRUE, message=TRUE}
#ohio parks
oh_parks <- sf::read_sf("./static_mapping/oh_parks.gpkg") #geopackage

#check crs
st_crs(oh_parks) == st_crs(sa) 

#reproject ohio parks
oh_parks <- st_transform(oh_parks, st_crs(sa))
st_crs(oh_parks) == st_crs(sa) #yippee they match

#filtering to just the study area
parks_sa <- sf::st_intersection(sa, oh_parks)

#plot to check
plot(parks_sa) #it worked, but there are many geometries

#tmap time
parks_map <-tm_shape(parks_sa) + 
              tm_polygons(fill="FEATTYPE", palette="brewer.greens",
                          title="Area Type") +
              tm_borders(col="black", lwd=2)


parks_map
```

### 2.4 — Water features and park intersections

The linear water data for the two counties are stored in separate directories:

- `../data/static_mapping/tl_2022_39133_linearwater/` (Portage County)
- `../data/static_mapping/tl_2023_39153_linearwater/` (Summit County)

Load both shapefiles, combine them using `dplyr::bind_rows()`, and reproject to match `sa`.

Next, determine which stream segments **spatially intersect** a park polygon. Use `sf::st_intersects()` to identify these segments and add a new column to your water features data frame (e.g., `intersects_park`) with a value of `TRUE` or `FALSE`. Use this column to symbolize the streams differently.

```{r water, echo=TRUE, message=TRUE}
p_water <- sf::read_sf("./static_mapping/tl_2022_39133_linearwater/tl_2022_39133_linearwater.shp") 
s_water <- sf::read_sf("./static_mapping/tl_2023_39153_linearwater/tl_2023_39153_linearwater.shp")

#check crs
st_crs(p_water) == st_crs(s_water) 

#bind
ps_water <- dplyr::bind_rows(p_water, s_water)

#fix crs
ps_water <- st_transform(ps_water, st_crs(sa))
#check crs
st_crs(ps_water) == st_crs(sa) 

#water intersects with park
water_parks <- ps_water %>% 
  mutate(intersects_park = lengths(st_intersects(geometry, parks_sa)) >0 )

# Filtering for only rivers that intersect parks
rivers_in_parks <- water_parks %>% filter(intersects_park == TRUE)

plot(rivers_in_parks)

water_map <- tm_shape(rivers_in_parks) + 
              tm_lines(col="#0019FF", lwd=2)
water_map
```

### 2.5 — Assemble the local map

Combine your places, parks, and water layers into a single map variable called `map_local`.

```{r map_local, echo=TRUE, message=TRUE}
# map map map
map_local <- places_map_nolab + parks_map + water_map +
        tm_shape(places_sa) + 
              tm_text("NAME.1", 
                     options = opt_tm_text(shadow=TRUE, remove_overlap = TRUE),
                     size = 0.9,  fontface = "bold") +
        tm_layout(title = "Parks and Rivers in Summit and Portage County",
                  title.position = tm_pos_out("center", "top"))

map_local
```

**Reflection question 2 (answer in your repo):** Describe the cartographic decisions you made about visual hierarchy in this map. Which features should visually dominate, and what specific symbol choices (size, color, transparency) did you make to achieve that? Why?
- This map was a bit challenging because there are many interesting and important features that need to be displayed with this map. I think the dominant feature that needs to be displayed is where parks are intersecting rivers/stream segments. To highlight this feature, I showed this area with a bright blue, thick line against the very pale yellow background showing just the summit and portage counties. Having the thicker line widths for the rivers helps to show their reaches when there are many boundaries and names of counties overlapping. The label for the cities was difficult to see in regular text settings so I assigned it a shadow.


---

## Task 3: Raster integration and layout elements

*Reference: Lovelace Ch 9.2.5 (layouts and map elements)*

Group 3 was responsible for integrating the DEM, adding layout elements, and combining everything. Let's start with the raster component.

### 3.1 — Prepare the DEM

Load `neoh_dem.tif` using `terra::rast()`.

Before cropping, **project** it to match the CRS of `sa` using `terra::project()`. Then crop to the study area extent. Add a comment explaining your workflow choices.

```{r dem, echo=TRUE, message=TRUE}
# Load in DEM
neoh_dem <- terra::rast("./static_mapping/neoh_dem.tif") 

# project dem, crop, and then mask to the study area
dem <- neoh_dem %>%  #take raster in 
  terra::project(., sa) %>% #project it to the same as sa
  terra::crop(., sa) %>% #crop it to sa bbox
  terra::mask(., sa) #mask to actual boundaries

plot(dem) #plot to check result

#not too large of a raster so I will project it first before cropping

```

*Tip: Projecting a large raster can be slow. Cropping AFTER projecting (as shown here) crops the projected version; cropping BEFORE projecting (to a smaller area first) can be faster. There are trade-offs — consider what's appropriate here.*

### 3.2 — Build the integrated local map

Now build a more complete version of the local map that incorporates the DEM. Requirements:

- The DEM as the base layer, using a continuous color scale appropriate for elevation (e.g., `YlGn`, `terrain`, or another sequential palette)
- The Portage and Summit county polygons overlaid with **semi-transparency** (`fill_alpha`) so the elevation shows through
- Your parks and water features from Task 2
- A north arrow: `tm_compass()`
- A scale bar

Store this as `map_local_elev`.

```{r map_local_elev, echo=TRUE, message=TRUE}
# make DEM tmap
oh_dem <- tm_shape(dem) +
  tm_raster(col.scale = tm_scale_continuous(values = "terrain"),
            col.legend = tm_legend(position = tm_pos_out("center", "bottom"),
                                   orientation = "landscape",
                                    title="Elevation")) +
  tm_shape(sa) + 
    tm_fill(fill_alpha = 0.2) +
    tm_borders(col = "black")

oh_dem

#combining the map
map_local_elev <- oh_dem +
      tm_shape(sa) + 
        tm_fill(fill_alpha = 0.2) +
        tm_borders("black", lwd=2) +
  parks_map +
  water_map +
  tm_compass(type = "8star", position = c("right", "bottom")) +
  tm_scalebar(position = c("right", "bottom")) +
  tm_title_out("Parks and Intersecting Rivers in Portage and Summit Counties")
  

map_local_elev
```

**Reflection question 3 (answer in your repo):** How does adding the elevation layer change the map's visual story? Did it create any visual hierarchy problems, and if so, how did you address them?
-It changes the story because it adds another feature and even more information about the river segements and park boundaries by explaining how high they are and what direction they could be possibly flowing. I started off with a more green palette that made it very difficult to see the parks, but I switched to the terrain palette to keep it a little less cluttered.
---

## Task 4: Multi-panel arrangement and inset maps

*Reference: Lovelace Ch 9.2.6 (faceted maps) and Ch 9.2.7 (inset maps)*

This task has two parts. The first uses `tmap_arrange()` — a straightforward approach to multi-panel maps. The second builds a proper inset, which is more powerful but more involved.

### 4.1 — Side-by-side with `tmap_arrange()`

Use `tmap_arrange()` to display `map_ohio` and `map_local_elev` side by side. Experiment with the `ncol` and `asp` arguments to get a reasonable layout.

```{r tmap_arrange, echo=TRUE, message=TRUE}
#new dem to fix the legend stacking issue
oh_dem_new <- tm_shape(dem) +
  tm_raster(col.scale = tm_scale_continuous(values = "terrain"),
            col.legend = tm_legend(position = tm_pos_out("center", "bottom"),
                                    title="Elevation")) +
  tm_shape(sa) + 
    tm_fill(fill_alpha = 0.2) +
    tm_borders(col = "black")

oh_dem_new

#recombining with the different legend set up now
map_local_elev_new <- oh_dem_new +
      tm_shape(sa) + 
        tm_fill(fill_alpha = 0.2) +
        tm_borders("black", lwd=2) +
  parks_map +
  water_map +
  tm_compass(type = "8star", position = c("right", "bottom")) +
  tm_scalebar(position = c("right", "bottom")) +
  tm_title_out("Parks and Intersecting Rivers in Portage and Summit Counties")
  
#quick check
map_local_elev_new

#changing their margins so they fit together
map_ohio1 <- map_ohio +
  tm_layout(meta.margins = c(0.3, 0, 0.02, 0)) + #bottom, left, top, right
  tm_title_out("Female Population Proportion in Ohio Counties") 

map_local_elev1 <- map_local_elev_new +
  tm_layout(meta.margins = c(0.3, 0, 0.02, 0))

#new tmap arrange
tmap_arrange(map_ohio1, map_local_elev1) #can't seem to stop those legends from overlapping

#also can't figure out why the titles are so small... I keep messing with it but no luck
```

**Reflection question 4 (answer in your repo):** `tmap_arrange()` is easy to use, but it has real limitations compared to a true inset. What are at least two limitations? (Think about how the two maps relate spatially and what information is or isn't communicated.)
-Some limitations come with the maps appearing to be the same size when one is the entirety of ohio and the other is two small counties from that map. Another limitation is having the legends not overlapping, in weird spots, or changing when the maps need to be plotted. Another situation is that it can be hard to tell which one is the guiding map because you have to have them vertically or horizontally stacked, when you may want to focus on some information, you may have to sacrifice the plotting order to have context.

### 4.2 — Building the inset map

A true inset map places a small context map *inside* the main map — a common technique for showing where a study area sits within a larger geographic context. Achieving this in `tmap` requires the `grid` package's viewport system.

Below is a partially complete code skeleton. **Read every comment carefully** before filling in the blanks. The concepts behind the viewports are more important than getting the aesthetics perfect on the first try.

```{r inset_map, echo=TRUE, message=TRUE}

# ---- Step 1: Build the inset (context) map ----
# This is the small map of all of Ohio that shows where our study area is.
# oh_region is the bounding box polygon you created in Task 2.1.

map_inset <- tm_shape(allohio) + tm_polygons() +        # fill in: all Ohio counties
  tm_shape(oh_region) + tm_borders(lwd = 2, col = "red") +  # fill in: border width
  tm_layout(bg.color = "lightblue") #+
  tm_scalebar(position = c("left", "bottom"))



# ---- Step 2: Normalize dimensions ----
# A viewport is defined in "normalized parent coordinates" (npc) from 0 to 1.
# If we just use width and height directly, maps with very different aspect ratios
# will appear distorted. norm_dim() calculates the width and height as fractions
# of the longest dimension, so that the aspect ratio is preserved.

norm_dim <- function(obj) {
  bbox <- st_bbox(obj)
  width  <- bbox[["xmax"]] - bbox[["xmin"]]
  height <- bbox[["ymax"]] - bbox[["ymin"]]
  w <- width  / max(width, height)
  h <- height / max(width, height)
  return(unit(c(w, h), "snpc"))
}

# Calculate dimensions for both maps
# The main map is based on the study area bounding box (oh_region)
# The inset map is based on all of Ohio
main_dim <- norm_dim(oh_region)  # fill in: use oh_region
ins_dim  <- norm_dim(allohio)  # fill in: use the all-Ohio sf object


# ---- Step 3: Define viewport positions ----
# A viewport is a rectangular region of a graphics device.
# main_vp defines the size of the main map within the graphics window.
# ins_vp defines the size AND position of the inset.

main_vp <- viewport(width = main_dim[1] * 1.8, height = main_dim[2] * 1.75)

ins_vp <- viewport(
  width  = ins_dim[1] * 0.4,              # scale the inset to 40% of normalized Ohio width
  height = ins_dim[2] * 0.4,              # scale proportionally
  x      = unit(1, "npc") - unit(4, "cm"),  # right edge of device minus a small margin
  y      = unit(0.01, "npc") + unit(1, "cm"),             # near the bottom of the device
  just   = c("right", "bottom")           # anchor point for the inset
)


# ---- Step 4: Draw ----
# grid.newpage() clears the device.
# print(..., vp = ...) draws a tmap within a specific viewport.
# pushViewport() makes the main viewport the "active" coordinate system,
#   so the inset is positioned relative to the main map, not the whole device.

grid.newpage()
print(map_local_elev1, vp = main_vp)
pushViewport(main_vp)
print(map_inset, vp = ins_vp)
```

Your first attempt will not look perfect — that is expected and intentional. You will need to:

- Fill in the four blanks (`___`)
- Adjust the scale factor (`0.4`) and margin (`0.5, "cm"`) to improve the inset's size and placement
- Add a title using `tm_title()` or `tm_layout(title = ...)` to `map_local_elev`

*Note: Getting the position right requires iterating. Try different values for the scale factor (e.g., `0.3`, `0.5`) and the margin. A scale factor around `0.35–0.45` and bottom-right or bottom-left placement usually works well for this study area.*

**Reflection question 5 (answer in your repo):** In your own words, explain what `norm_dim()` is doing and why it is necessary. What specific visual problem would occur if you skipped that step and passed the raw coordinate extents to `viewport()` directly?
- norm_dim() takes a bounding box of this area sf object and turns it into a width and height with normalized proportions so that it can fit within a viewport basically. Without this function, the viewport wouldn't be created because it can't take geographic points and turn it into a width and height later on. I think it would end up having a lot of issues and fail essentially in the end because it couldn't scale up.

---

## Task 5: Your own static map

Using whatever spatial data, extent, and domain you choose, create an original static map. Requirements:

1. It must include **vector AND raster data** used together in a meaningful way
2. It must include spatial data representing a **social process** (e.g., political boundaries, land use, population) AND an **environmental process** (e.g., elevation, water, vegetation, climate)
3. The map should **stand on its own** and communicate its purpose without additional explanatory text — if a reader needs to ask "what am I looking at?", the map hasn't done its job
4. Good coding practices: comments, relative paths, clean variable names

You may use data from any of the course datasets (CBW, Ohio, Lake Erie) or data you obtain independently. If using external data, document the source in a comment.

```{r your_map, echo=TRUE, message=TRUE}
# One of the undergrads in my lab group collected data at the same wetland I did, I will be using some of her SURE work and data points for this exercise
emily_data <- read_csv("./EmilyCampbell_Combined_SURE_Work.csv")
#turning her work in an sf object
emily_data_sf <- st_as_sf(emily_data, coords = c("longitude", "latitude"), crs = "sa")

#NE Ohio national landuse cover data
neoh_nlcd <- terra::rast("./neoh_nlcd.tif") 

#DEM of the wetland area (TRUC aka Trumbull Creek) itself
TRUC_dem <- terra::rast("./dem_partially_filled.tif")

# Study area of project
boundary <- sf::read_sf("./PROJECT_LIMITS.shp") 

#checking crs of these rasters
st_crs(neoh_nlcd)
st_crs(TRUC_dem)

#reprojecting boundary to the same as SA from earlier
boundary_new <- st_transform(boundary, st_crs(sa))

TRUC_dem_new <- TRUC_dem %>%  #take raster in 
  terra::project(., boundary_new) %>% #project it to the same as sa
  terra::crop(., boundary_new) %>% #crop it to boundary bbox
  terra::mask(., boundary_new)  #mask to actual boundaries

neoh_nlcd_new <- neoh_nlcd %>%  #take raster in 
  terra::project(., boundary_new, method = "near") %>% #project it to the same as sa
  #Had to do this because it is resampling and has decimal values for nlcd
  terra::crop(., boundary_new) %>% #crop it to boundary bbox
  terra::mask(., boundary_new) #mask to actual boundaries

#getting the real nlcd color palette
nlcd_cols <- c(
  "11" = "#476BA1",  # Open Water
  "21" = "#DDC9C9",  # Developed, Open Space
  "22" = "#D89382",  # Developed, Low Intensity
  "23" = "#ED0000",  # Developed, Medium Intensity
  "24" = "#AA0000",  # Developed, High Intensity
  "31" = "#B2ADA3",  # Barren
  "41" = "#68AA63",  # Deciduous Forest
  "42" = "#1C6330",  # Evergreen Forest
  "43" = "#B5C98E",  # Mixed Forest
  "52" = "#CCBA7C",  # Shrub/Scrub
  "71" = "#E2E2C1",  # Grassland/Herbaceous
  "81" = "#DED83E",  # Pasture/Hay
  "82" = "#AA7028",  # Cultivated Crops
  "90" = "#BAD8EA",  # Woody Wetlands
  "95" = "#70A3BA")   # Emergent Herbaceous Wetlands

#had issues with the categorical raster, now I am converting it to a factor first
# Convert NLCD raster to categorical
neoh_nlcd_new <- terra::as.factor(neoh_nlcd_new)


#creating a map showing the dem of TRUC (environmental), the landuse designation (social), and emily's data points(vector)

TRUC_map <-
  tm_shape(TRUC_dem_new) + #raster dem of TRUC
    tm_raster(col.scale = tm_scale_continuous(values = "terrain"),
              col.legend = tm_legend(position = tm_pos_out("center", "bottom"),
                                     title="Elevation",
                                     orientation="landscape")) + #legend on the bottom
  tm_shape(boundary_new) +  #TRUC shapefile parcel boundary
    tm_fill(fill_alpha = 0.2) +
    tm_borders(col = "black") +
  tm_shape(neoh_nlcd_new) +
    tm_raster( #for some reason the colors are breaking and are not plotting on here :(
          col.scale = tm_scale_categorical(values = nlcd_cols), #categorical data and my color palette
          col_alpha = 0.6, #slightly transparent
          title = "Historic NLCD Land Cover Classes") +
      tm_scalebar(position = c("center", "bottom")) +
      tm_compass(position = c("right", "top"), size = 1.5) +
  tm_shape(emily_data_sf) +
    tm_dots(size="P_mg_kgsoil", fill="gold",
            size.legend = tm_legend("Measured Phosphorus per mg/kg of Soil")) +
  tm_layout(#I can't seem to get this title any bigger, tried both tm_title and tm_layout
      main.title = "Soil Phosphorus Measured Throughout the Newly Restored
Trumbull Creek Wetland and its Previous National Land Cover Class Designation",
      main.title.size =0.7, #any higher shoots the title away from the map
      main.title.position = "center",
      main.title.fontface = "bold")

TRUC_map 
```

- Found the nlcd color palette here https://smalltownbigdata.github.io/feb2021-landcover/feb2021-landcover.html
- I can't get the right colors to work and the title is a little small. I also couldn't figure out giving NLCD a proper label other than the numbers

---

## Questions

1. Describe and justify your cartographic choices in Task 1 (Ohio choropleth). Specifically address: your classification method, your color palette, and how these choices serve the map's intended communication goal. Would the map tell a different story with different choices?

2. In Task 2.4 you identified stream segments that intersect parks. Describe your approach step by step, as if explaining it to a classmate who hasn't done spatial data analysis before. Why is this distinction cartographically meaningful (*i.e., why might a map reader care which streams run through parks)?

3. In Task 4, you used both `tmap_arrange()` and the `grid` viewport approach to create multi-panel maps. What are the trade-offs between these two methods? In what professional or analytical context would you prefer one over the other?

4. Compare the cartographic decisions you made individually in this lab to the decisions your group made collectively during the in-class frankenmap exercise. Where did your choices diverge, and what explains the difference?

5. Describe your original map in Task 5: what spatial story are you trying to tell, did you achieve your communication goal, and what would you change given more time?

