---
title: "in-class dynamic mapping"
author: "Patrick Bitterman"
date: "2026-04-07"
output: html_document
---

# Today's schedule

- Open discussion
- Intro to Leaflet (https://rstudio.github.io/leaflet/)



Packages

```{r}
library(tidyverse)
library(terra)
library(sf)
library(leaflet) #new
library(leaflet.extras) #new
```

# Leaflet

Leaflet is....

...a Javascript library with an API we can access in R. 

Wait, that's a LOT of jargon. Can we add some clarity?

- **JavaScript**: programming language that runs inside web browsers. It powers interactive behavior on almost every website you use
- **Library**: collection of pre-written code that solves a specific problem. Leaflet is a JS library built specifically for interactive maps — someone else wrote the hard parts, and we use the result
- **API (Application Programming Interface)**:  set of functions and rules that lets one piece of software talk to another. Leaflet's API is the list of commands you can give it: "add a tile layer," "add a marker," etc.
- **Access in R**: the `leaflet` R package is a *wrapper*: it translates your R code into JavaScript instructions that Leaflet understands, so you never have to write JavaScript yourself.

```{r}

```

## Features of Leaflet
- Interactive panning/zooming
- Compose maps using arbitrary combinations of:
    - Map tiles
    - Markers
    - Polygons
    - Lines
    - Popups
    - GeoJSON
    
## Functionality
- Create maps right from the R console or RStudio
- Embed maps in knitr/R Markdown documents and Shiny apps
- Easily render spatial objects from the sp or sf packages, or data frames with latitude/longitude columns
- Use map bounds and mouse events to drive Shiny logic
- Display maps in non spherical mercator projections
- Augment map features using chosen plugins from leaflet plugins repository

# A quick detour: Functional vs. Object-Oriented Programming

Before we dive into examples, notice *how* Leaflet is written in R, as it reveals something important about programming paradigms.

**Object-Oriented Programming (OOP)** organizes code around *objects* that bundle data and behavior together. You call methods on an object, which often mutates its state. The underlying JavaScript Leaflet library is OOP:

```javascript
// JavaScript (OOP style)
var map = L.map('mapid');        // create an object
map.setView([41.15, -81.35], 14); // call a method on it — mutates the object
L.tileLayer('https://...').addTo(map); // another object added to the map
```

**Functional Programming (FP)** organizes code around *functions* that take inputs and return outputs — no hidden state changes. You compose functions together, passing the result of one as input to the next. The R leaflet package expresses the same idea in a functional style:

```r
# R (functional style)
leaflet() %>%        # function returns a map object
  setView(...) %>%   # function takes that object, returns a new one
  addTiles()         # function takes that, returns a new one
```

Each `%>%` step is a function call. Nothing is mutated — each call produces a new value. This is the same philosophy as `dplyr`: `filter() %>% select() %>% mutate()`.

**Key insight:** The R leaflet package wraps an OOP JavaScript library in a functional R interface. You build maps by composing functions, but the *result* is an object (you'll see an example later when we use `$` to inspect it).

---

# Demonstration

A simple setup

```{r}
m <- leaflet()

m
```

What do you get?


Let's try again

```{r}
m <- leaflet() %>%
  addTiles()

m
```

And now? What do you see?


Let's build on that

```{r}
m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng = -81.350903, lat = 41.150377, popup="Our McGilvrey Hall GIS Lab")
    #could pass it as a series of data as well
m
```

What about now?

## Let's add some complexity

Create some data to plot. Let's break this down a bit. What am I trying to do?

```{r}

# First let's define a color palette we can sample from - this is only really necessary for the demo
mypal <- RColorBrewer::brewer.pal(12, "Set3") #making your own palette, 12 colors from set 3

# start with a data frame
df <- data.frame(
  lat = rnorm(100) * 2 + 41, #creating a vector of 100 values with a normal distribution 
  lng = rnorm(100) * 2 - 81.5, 
  size = runif(100, 5, 20), #creates random deviates, number, min, max
  color = sample(mypal, 100, replace = T) #assigning our custom palette
)

      # rnorm starts on 0, 1 as std
      # he is pulling 100, multiplying by 2, and centering around our area
      # runif, sampling from a uniform distribution (equal odds) in upper and lower tails
      # We have only created the data, we didn't set it on the map

# then add the data frame to a leaflet map
m2 <- leaflet(df) %>% addTiles()

m2
```


What do you get? Why?

How can we interrogate the properties/attributes of an object?

The `$` operator

```{r}
m2$x #we do this to look at what's inside
```

What do the data look like?

> **OOP note:** Even though we *built* `m2` by composing functions (`leaflet() %>% addTiles()`), the result is an R object (an S3 list) with named properties. The `$` operator is how you access those properties, and it's the same way you'd access fields on an object in OOP. So we construct using functional paradigms, but get an OOP result.


## Back to the show

Let's try again to visualize it. Again, break down the code first

```{r}
# first one
m2 %>% addCircleMarkers(radius = ~size, color = ~color, fill = FALSE)
  # use ~ notation to assign the column name

# second one
m2 %>% addCircleMarkers(radius = runif(100, 4, 10), color = 'red') 
  # only uses the color red, This makes new data instead of using the old data

```

What happened this time???

Let's add some more tiles

```{r}
# Let's check out some other tiles

m <- leaflet() %>% setView(lng = -81.34, lat = 41.145, zoom = 14)
m %>% addTiles()


# third party tiles using addProvider() function

m %>% addProviderTiles(providers$Stadia.StamenToner) #black and white, diff fonts
m %>% addProviderTiles(providers$CartoDB.Positron) # boring
m %>% addProviderTiles(providers$CartoDB.DarkMatter) #all dark
m %>% addProviderTiles(providers$Esri.NatGeoWorldMap) #has elevation data

#can find more in the documentation for provider tiles
```

Give it a shot. What do you like?


## Let's use some of our data

```{r}
parks <- sf::read_sf("./data/static_mapping/oh_parks.gpkg") %>%
  sf::st_transform(., "EPSG:4326") # transform to WGS84 to make Leaflet happy

# set up the map, zoom out a bit
mp <- leaflet(data = parks) %>% setView(lng = -81.34, lat = 41.145, zoom = 10)
mp %>% addTiles() %>% 
  addPolygons(popup = ~NAME, label = ~NAME) #same idea as tmap, label shows up for a mouse


```

What's the difference between popup and label?



## Lines

Who wants to help break down this code?

```{r}
portage_streams <- sf::read_sf("./data/static_mapping/tl_2022_39133_linearwater/tl_2022_39133_linearwater.shp") %>%
  sf::st_transform(., "EPSG:4326") # transform to WGS84 to make Leaflet happy

leaflet(data = portage_streams) %>% 
  setView(lng = -81.34, lat = 41.145, zoom = 10) %>% 
  addTiles() %>%
  addPolylines(., color = "blue", #because these are lines
               popup = ~paste0(FULLNAME, ": ", LINEARID)) #creating new string
              #paste0 puts the strings: it's ID

```

What happened?



## Multiple layers

(Note, there's a difference here in calling various datasets)

```{r}
m.both <- leaflet() %>% # not using the dataset as a parameter here - why not???
  setView(lng = -81.34, lat = 41.145, zoom = 10) %>% 
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>%
  addPolygons(data = parks, popup = ~NAME, label = ~NAME, color = "green") %>% #first layer
  addPolylines(data = portage_streams, color = "blue", #second layer
               popup = ~paste0(FULLNAME, ": ", LINEARID))

m.both #multilayer map

```


# Let's practice

- Add the municipal boundaries in Portage and Summit counties to this map (recall how we made our "study area" last week - you'll have to reuse last week's data)
- Give them a "popup" corresponding to their name
- And give the parks a color based on the TYPE of park.



# Team activity

- Everyone has reviewed the readings/walkthrough for this week (https://rstudio.github.io/leaflet/)
- Today, we will...
  1. break into teams
  2. complete tasks using Leaflet
  3. present (informally) your results, methods, challenges, etc. (maybe on Thursday)
  
- ALL DATA ARE SOMEWHERE IN THE REPOSITORY
- All tasks are included in the various readings or documentation
- For some of the extra challenges, you may need to look at the `leaflet.extras` package
  
## Tasks

### GROUP 1

- Make a choropleth map of the population density of Ohio Counties. There should be no more than 7 classes
- When the user "mouses over" a polygon, a pop-up with the county name and population density should appear
- Place a legend in the bottom-right corner

#### Extra challenges: 
1. Add a minimap to the interface
2. Add a "locate me" button to the interface


### GROUP 2

- Make a map of the monitoring stations and removed dams in the Chesapeake Bay Watershed
- The monitoring stations should be symbolized as a filled, semi-transparent circle with the size scaled to the Drainage Area of the station
- Dam layer should be shaded by the year it was removed
- Add controls to turn individual layers on and off

#### Extra challenges:
1. Use a custom icon for the dam layer
2. Add a control to the interface to change the base map


### GROUP 3

- Using the Ohio county data and a variable of your choice from the BMP or demographic datasets, compute local Moran's I (LISA) quadrant assignments (you may need to review the activities from that week)
- Make a choropleth map shaded by LISA quadrant (HH, HL, LH, LL) using a meaningful four-color palette
- When the user clicks a county, a pop-up should display the county name and the local Moran p-value
- Add a legend

#### Extra challenges:
1. Add a layer toggle control so the user can switch between the LISA quadrant map and the underlying variable choropleth
2. Add a `setView()` call with a reasonable default zoom centered on Ohio

amindianat/poptot proportion
queen
```{r}
#choropleth map creation
choropleth <- 
  tm_shape(county_census) + 
    tm_polygons(fill = "prop_amindianat", 
                fill.scale = tm_scale_intervals(style = "jenks", 
                                                values = "brewer.purples", n = 5),
                fill.legend = tm_legend(title = "Proportion of American Indians 
      and Native Americans")) + 
    tm_borders(lwd = 1.5, col = "black") 

#check choropleth map of American Indian/ Native American Population Proportion
choropleth

```



```{r}

library(tidyverse)
library(terra)
library(sf)
library(leaflet) #new
library(leaflet.extras) #new
library(spdep)
library(qualpalr)
library(tmap)

#read and join data
oh_counties <- sf::read_sf("./data/static_mapping/oh_counties.gpkg")
oh_census <- sf::read_sf("./data/static_mapping/oh_counties_DP2020.csv")

county_join <-left_join(oh_counties,oh_census, join_by("GEOIDFQ" == "geoid"))

#looking at native american population
# doing a proportion of total population
county_census <- county_join |> 
  mutate(prop_amindianat = as.numeric(amindianat)/ as.numeric(poptotal))

## Moran's I -LISA for native american population

#Look at our data with a histogram and by county
am.pop.counties <- county_census |> 
    ggplot(aes(x = NAME, y = prop_amindianat)) +
        geom_col(fill="steelblue") + 
    labs(x="County Name",
         y="Proportion of Population Per County",
         title="Proportion of Native American Population in Ohio Counties") +
        theme_minimal()

am.pop.counties #very small proportions per county

#check crs first
sf::st_crs(county_census) 


# Build spatial weights matrix
coords <- sf::st_geometry(county_census)
nb  <- spdep::poly2nb(county_census, queen = TRUE)
#average number of links = 5.25, it accounts for all counties

#row standardize the weights
lw  <- spdep::nb2listw(nb, style = "W", zero.policy = TRUE)

# Compute Local Moran's I
lisa <- spdep::localmoran(county_census$prop_amindianat, lw, zero.policy = TRUE)
county_census$local_I  <- lisa[, "Ii"] #moran's i column
county_census$lisa_pval <- lisa[, "Pr(z != E(Ii))"] #moran's pvalue

#LISA Quadrant Labels
# Mean-center the variable and the spatial lag
x     <- county_census$prop_amindianat
x_lag <- lag.listw(lw, x, zero.policy = TRUE)

x_mean     <- mean(x,     na.rm = TRUE)
xlag_mean  <- mean(x_lag, na.rm = TRUE)


# create the quadrants
county_census <- county_census %>%
  mutate(
    quad = case_when(
      x >  x_mean & x_lag >  xlag_mean ~ "HH",
      x >  x_mean & x_lag <= xlag_mean ~ "HL",
      x <= x_mean & x_lag >  xlag_mean ~ "LH",
      x <= x_mean & x_lag <= xlag_mean ~ "LL"
    ),
    # Mark non-significant as NS (p > 0.05)
    quad_sig = ifelse(lisa_pval > 0.05, "NS", quad))


#creating the color scheme for the LISA
county_census <- county_census |>
  mutate(color =  case_when( #use this later to classify them
                              quad_sig == "HH" ~ "#d7191c",
                              quad_sig == "HL" ~ "#fdae61",
                              quad_sig == "LH" ~ "#abd9e9",
                              quad_sig == "LL" ~ "#2c7bb6",
                              quad_sig == "NS" ~ "#d3d3d3")) #not significant p-values are grey


#Popup to get the moran's i and the p-value
county_census <- county_census %>%
  mutate(
    popup_lisa = paste0(
      "<b>", NAME, " County</b><br>",
      "LISA Quadrant: <b>", quad_sig, "</b><br>",
      "Local Moran p-value: <b>", round(lisa_pval, 4), "</b>" ))

#color palette for the leaflet - choropleth, not LISA
#using the quantile method
pal_purple <- colorQuantile(
  palette = "Purples",
  domain  = county_census$prop_amindianat,
  n    = 4)    # adjust number of bins as you like



#Reproject for Leaflet
county_wgs <- st_transform(county_census, crs = 4326)

#Map
leaflet(county_wgs) %>%
  addMapPane("choropleth", zIndex = 410) %>% #ranking which one comes first
  addMapPane("lisa",     zIndex = 420) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap,
                   group = "NatGeo World Map")  %>% #base map 1
  addProviderTiles(providers$Esri.WorldImagery,
                   group = "Satellite View")  %>% #base map 2
  setView(lng = -81.34, lat = 41.145, zoom = 6) %>% #SetView window and zoom
  addPolygons( #start with choropleth map
    group = "Choropleth Map",
      fillColor = ~pal_purple(prop_amindianat),#purple palette and native american proportion
      fillOpacity = 0.9,
      color = "black",    
      weight = 1,
      options = pathOptions(pane = "choropleth")
  ) %>%
  addPolygons( #LISA quadrant map
    group= "LISA Quadrant",
    fillColor   = ~color,
    fillOpacity = 0.9,
    color       = "black",
    weight      = 2,
    popup       = ~popup_lisa,
    highlightOptions = highlightOptions(
      weight       = 2,
      fillOpacity  = 0.9,
      bringToFront = TRUE
    )) %>%
  addLegend(
  position = "bottomright", #stacking legends
  title    = "LISA Quadrant",
  colors   = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6", "#d3d3d3"), #rewriting, had an issue with length of variable
  labels   = c("HH – High-High (cluster)",
               "HL – High-Low (outlier)",
               "LH – Low-High (outlier)",
               "LL – Low-Low (cluster)",
               "NS – Not Significant"),
  opacity  = 0.9) %>%
  addLegend( #legend for the other choropleth
  position = "bottomright",
  title    = "Native American Population Proportions",
  pal   = pal_purple,
  values   = ~prop_amindianat,
  opacity  = 1) %>%
  addLayersControl(
    baseGroups = c("NatGeo World Map", "Satellite View"), 
    overlayGroups = c("Choropleth Map", "LISA Quadrant"), 
    options = layersControlOptions(collapsed = FALSE))


```
https://r-statistics.co/Interactive-Maps-in-R-with-leaflet.html
https://rstudio.github.io/leaflet/articles/colors.html#continuous-input-discrete-colors-colorbin-and-colorquantile

