---
title: "Lab 06"
#author: "Dr. Patrick Bitterman"
#date: "`r format(Sys.time(), '%Y %B %d')`"
output: html_document
my_subtitle: "Env Analysis in R"
geometry: margin=1in
---

# Lab 06: Dynamic mapping with Leaflet

### Read the instructions COMPLETELY before starting the lab

This lab builds directly on labs 2–4 and the in-class Leaflet exercises. You will need those lab files available for reference.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

*Note: This lab is output as an HTML document — not a PDF. Leaflet maps are **interactive** and cannot be embedded in a static PDF. Your knitted output will be an `.html` file. Keep this in mind when organizing your submission.*

## Introduction

In the previous lab you built static maps using `tmap`. Static maps are excellent for print, publication, and controlled presentation. But many modern mapping applications demand **interactivity**: users want to pan, zoom, click, and toggle layers. Leaflet is the dominant open-source library for interactive web maps, and `leaflet` (the R package) wraps it for use in R.

Leaflet's design philosophy is similar to `tmap` in that maps are built by **layering components**:

```
leaflet() %>%
  addTiles() %>%                  # base map (tiles)
  addPolygons(...) %>%            # vector layer
  addLegend(...) %>%              # legend
  addLayersControl(...)           # layer toggle
```

Each `add*()` call adds a new layer or control to the map object. The `%>%` pipe passes the map forward. This means you can build up a map incrementally, verify each piece, and compose them together — exactly as you did in `tmap`.

Key resources:
- Official leaflet for R docs: https://rstudio.github.io/leaflet/
- Making maps with R (Hahn, Ch. 5): https://bookdown.org/nicohahn/making_maps_with_r5/docs/leaflet.html
- `leaflet.providers` for basemap tile options

```{r packages, echo=TRUE, message=TRUE}
library(sf)
library(leaflet)
library(tidyverse)
library(spdep)   # needed for Task 2 (Moran/LISA)
library(leaflet.extras) #extra functionality
library(raster) # to use raster images in leaflet
library(dplyr)
```

*Note: You may need additional packages depending on your Task 3 choices. Load them here.*

---

## Task 1: Choropleth from Lab 2

*Reference: leaflet docs — `addPolygons()`, `colorBin()`, `addLegend()`*

In Lab 2, Task 2.3, you created a choropleth of Chesapeake Bay Watershed counties shaded by the total cost of BMPs funded/implemented in each county. Here you will recreate that map in Leaflet.

### 1.1 — Reload and prepare the data

Go back to your Lab 2 code and reload the county geometry and BMP cost data. Perform whatever join and preprocessing was necessary to produce your choropleth variable. Store the result as an `sf` object called `counties_bmp`.

Before mapping, **check and set the CRS**. Leaflet requires geographic coordinates (WGS 84, EPSG 4326). If your data are in a projected CRS, reproject now using `sf::st_transform(counties_bmp, 4326)`.

```{r task1_data, echo=TRUE, message=TRUE}
# Load in the county data set from CBW
counties <- sf::read_sf("./CBW/County_Boundaries.shp") %>% 
  sf::st_make_valid()

#bmp cost data
bmps <- read_csv("./CBW/BMPreport2016_landbmps.csv")

 #getting the county FIPS code pulled out of BMPs 
unique_bmps <- bmps |> mutate(COUNTYFP10 = stringr::str_sub(GeographyName, 3, 5), # matching columns
                       STATEFP10 = stringr::str_sub(GeographyName, 1, 2)) |>
                       group_by(STATEFP10, COUNTYFP10) |>  #need to group by both
                summarize(TotalCost = sum(Cost, na.rm=TRUE), .groups="drop") |> #total cost per county
                distinct() #keep only the unique state and counties

#unique counties
unique_counties <- counties |>
  dplyr::select(STATEFP10, COUNTYFP10, NAMELSAD10, geometry) |> 
  distinct() #keeping only the unique counties (not all entries)

#combine counties and BMPS by STATE and COUNTY FP10
counties_bmp <- unique_counties |> left_join(unique_bmps, by = c("COUNTYFP10","STATEFP10")) 

#check crs of the counties_bmp
sf::st_crs(counties_bmp) #looks all good, its in WGS 84

```

### 1.2 — Define a color palette

Leaflet's `colorBin()` function creates a color palette from a set of breaks. You must determine the **equal-interval breaks programmatically** — do not hard-code them.

*Hint: The range of the variable divided by the number of classes gives you the interval width. `seq()` can generate your break points.*

```{r task1_palette, echo=TRUE, message=TRUE}
#define the breaks
breaks <- seq(min(0), max(200000000), length.out=7) # break points, 7 seemed best
# has it broken down to this range representing millions

# palette
pal <- colorBin(palette = "BuPu", domain = 0:200000000, bins = breaks) #blue to purple palette
```

*Requirement: Do NOT use the default color scheme (Blues). Choose a palette appropriate for sequential data from a source like ColorBrewer.*

### 1.3 — Build the choropleth map

Build the Leaflet map with the following requirements:

- The county polygons colored by total BMP cost using your `colorBin()` palette
- A **mouse-over label** that displays the county name and total cost when the cursor hovers over a county. Use the `label` argument in `addPolygons()`.
- A **legend** using `addLegend()`
- A basemap tile layer using `addProviderTiles()` — choose something other than the default OpenStreetMap. Browse options at https://leaflet-extras.github.io/leaflet-providers/preview/

```{r task1_map, echo=TRUE, message=TRUE}
#Map it outttt
leaflet(counties_bmp) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap,
                   group = "NatGeo World Map")  %>% #base map 1
  addProviderTiles(providers$Esri.WorldImagery,
                   group = "Satellite View")  %>% #base map 2
  addPolygons( #start with choropleth map
    group = "BMP Map",
      fillColor = ~pal(TotalCost), #my pal and total cost
      fillOpacity = 0.9,
      color = "black",    
      weight = 1,
      label = ~paste0(NAMELSAD10, ": $", #prints the county name with a space after
                      formatC(TotalCost, format = "f", digits = 0, big.mark = ","))) %>%
                      #has the total cost after, no decimals, just commas
  addLegend( #legend for the  choropleth
  position = "bottomright",
  title    = "BMPs Funded/ Implemented in Each County",
  pal   = pal,
  values   = ~TotalCost,
  opacity  = 1,
  labFormat = labelFormat(prefix = "$", big.mark = ",")) %>% #fixes scientific notation
  addLayersControl(
    baseGroups = c("NatGeo World Map", "Satellite View"), 
    overlayGroups = ("BMP Map"), 
    options = layersControlOptions(collapsed = FALSE))

```

**Reflection question 1 (answer in your repo):** In `tmap`, classification method choices (jenks, quantile, etc.) are made when you call `tm_polygons()`. In Leaflet, you define the palette and breaks separately using `colorBin()`, `colorQuantile()`, or `colorNumeric()`. What are the trade-offs of this approach? Does separating the palette from the geometry layer give you more or less flexibility? Give a specific example.
- I think there are some tradeoffs by using these two methods, but in general I think separating the palette from the geometry layer could give you more flexibility. In scenarios when you need to have lots of control over the palettes or ranges, using the leaflet method allows you to preset different colors and breaks to the data which can be resued over time in different maps. This allows a consistent and steady color scheme and helps with publishable figures. Also, being able to control it for multiple different layers can help you keep your code tidy and functioning. On the other hand, it is very simple and quick to just toss in the method choice within the tmap settings, which allows for quick, easy plotting if the map is just a check for future work.

---

## Task 2: LISA map from Lab 3

*Reference: leaflet docs — `addPolygons()` popup, `addLayersControl()`, `addProviderTiles()`*

In Lab 3, Bonus Task 2, you produced a choropleth map of LISA (local Moran's I) quadrants — a categorical four-class map showing spatial clustering patterns (HH, HL, LH, LL). Here you will recreate that map in Leaflet with interactive enhancements.

### 2.1 — Reload and prepare the data

Go back to your Lab 3 code. Reload the spatial data and rerun the Moran/LISA analysis to recover the quadrant assignments and p-values. Store the result as an `sf` object called `counties_lisa`, reprojected to WGS 84 (EPSG 4326).

*Note: You need both the quadrant labels and the p-values — make sure both are columns in `counties_lisa` before proceeding.*

```{r task2_data, echo=TRUE, message=TRUE}
#choosing to look at female ages 22-24 population per county in those CO, UT, NM, KS, WY 
#reading in the exported table from that task
f.pop.states.1 <- sf::read_sf("./f.pop.states.gpkg") 

f.pop.states.1 <- f.pop.states.1 |>
  mutate(female_proportion = (B01001_034E/B01001_001E)) # female (22-24) proportion of total pop in each county

# Build spatial weights matrix
coords <- sf::st_geometry(f.pop.states.1)
nb  <- spdep::poly2nb(f.pop.states.1, queen = TRUE)

#row standardize the weights
lw  <- spdep::nb2listw(nb, style = "W", zero.policy = TRUE)

# Compute Local Moran's I
lisa <- spdep::localmoran(f.pop.states.1$female_proportion, lw, zero.policy = TRUE)
f.pop.states.1$local_I  <- lisa[, "Ii"] #moran's i column
f.pop.states.1$lisa_pval <- lisa[, "Pr(z != E(Ii))"] #moran's pvalue

#LISA Quadrant Labels
# Mean-center the variable and the spatial lag
x     <- f.pop.states.1$female_proportion
x_lag <- lag.listw(lw, x, zero.policy = TRUE)
x_mean     <- mean(x,     na.rm = TRUE)
xlag_mean  <- mean(x_lag, na.rm = TRUE)


# create the quadrants
counties_lisa <- f.pop.states.1 %>%
  mutate(
    quad = case_when(
      x >  x_mean & x_lag >  xlag_mean ~ "HH",
      x >  x_mean & x_lag <= xlag_mean ~ "HL",
      x <= x_mean & x_lag >  xlag_mean ~ "LH",
      x <= x_mean & x_lag <= xlag_mean ~ "LL"))


```

### 2.2 — Define a categorical color palette

LISA quadrants are a categorical variable (four classes). Use `colorFactor()` instead of `colorBin()`. Assign a distinct, meaningful color to each quadrant (HH, HL, LH, LL). A common convention is warm colors for High clusters and cool colors for Low clusters.

```{r task2_palette, echo=TRUE, message=TRUE}
# making palette for the lisa, not coding it into the df
pal_lisa <- colorFactor(palette = c("#d7191c","#fdae61","#abd9e9","#2c7bb6"), 
                        levels = c("HH", "HL", "LH", "LL"))
```

### 2.3 — Build the LISA map

Build the Leaflet map with the following requirements:

- County polygons colored by LISA quadrant using your `colorFactor()` palette
- A **pop-up window** (not a hover label) that displays the p-value for that county when clicked. Use the `popup` argument in `addPolygons()`. Format the p-value to a reasonable number of decimal places.
- A **basemap switcher** that allows the user to toggle between **3 different basemaps**. Use `addProviderTiles()` with `group` arguments and `addLayersControl()` with `baseGroups`.
- A legend

```{r task2_map, echo=TRUE, message=TRUE}

#Popup to get the moran's i and the p-value
counties_lisa <- counties_lisa %>%
  mutate(
    popup_lisa = paste0(
      "<b>", NAME, " </b><br>",
      "LISA Quadrant: <b>", quad, "</b><br>",
      "Local Moran p-value: <b>", round(lisa_pval, 4), "</b>" ))

#check the crs before plotting
st_crs(counties_lisa) # its the 1983 datum so it needs reprojected for leaflet

#transform to leaflet crs
counties_lisa <- st_transform(counties_lisa, crs = 4326)

#turn quad into factor 
counties_lisa$quad <- as.factor(counties_lisa$quad)

#Map
leaflet(counties_lisa) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap,
                   group = "NatGeo World Map")  %>% #base map 1
  addProviderTiles(providers$Esri.WorldImagery,
                   group = "Satellite View")  %>% #base map 2
  addProviderTiles(providers$Stadia.StamenWatercolor,
                   group = "Stamen Watercolor")  %>% #base map 3
  addPolygons( #LISA quadrant map
    group= "LISA Quadrant",
    fillColor   = ~pal_lisa(quad),
    fillOpacity = 0.9,
    color       = "black",
    weight      = 2,
    popup       = ~popup_lisa,
    highlightOptions = highlightOptions(
      weight       = 2,
      fillOpacity  = 0.9,
      bringToFront = TRUE
    )) %>%
  addLegend(
  position = "bottomright", #stacking legends
  title    = "LISA Quadrant",
  colors   = pal_lisa(unique(counties_lisa$quad)), #this makes it just appear once
  labels   = unique(counties_lisa$quad), #this makes so that it only prints the values once, not 5 times
  opacity  = 0.9) %>%
  addLayersControl(
    baseGroups = c("NatGeo World Map", "Satellite View", "Stamen Watercolor"), 
    overlayGroups = "LISA Quadrant", 
    options = layersControlOptions(collapsed = FALSE))
```

*Tip: The `addLayersControl()` function accepts `baseGroups` (radio buttons, one active at a time) and `overlayGroups` (checkboxes, multiple can be active). Basemaps should be `baseGroups`.*

**Reflection question 2 (answer in your repo):** Compare the static LISA map you made in Lab 3 to this interactive version. What does interactivity add? Are there things the static map communicates more clearly? Consider map reading purpose (exploration vs. communication) in your answer.
-I feel like this interactive map is better for the exploration of this data and how these five states are related to each other. Unless this was published to a website, it would be a lot more difficult to get people to see this map compared to a static map that could be published in a manuscript or something else. The fact that this leaflet map has popups makes it far more convenient for showing different spatial patterns without overflowing the map with county names, p-values, and what the clusters are. It is also helpful to toggle between different basemaps to see if other factors are influencing this behavior like placement of mountain ranges, distance to the coastline, surrounding states, etc.

---

## Task 3: Free-form map from Lab 5

*Reference: leaflet docs — `addLayersControl()` with overlayGroups, your choice of other features*

In Lab 5, Task 5, you created an original map combining vector and raster data, representing both a social and an environmental process. Here you will recreate that concept in Leaflet.

*Note: Leaflet handles raster data differently than `tmap`. The `addRasterImage()` function in the `leaflet` package can display projected rasters, but requires the `raster` package (not `terra`) or conversion. An alternative is to represent the "environmental process" through a different vector layer (e.g., watersheds, land cover categories as polygons) if raster integration proves difficult. Document your approach in a comment.*

### 3.1 — Reload and prepare your data

Reload the data you used in your original free-form map. Reproject all layers to WGS 84 (EPSG 4326). List the layers you are using in a comment.

```{r task3_data, echo=TRUE, message=TRUE}
# To remake the map from lab 5 I will be using some data from my wetland lab group
# layers I'm using:
#  - Landuse Landcover: Social layer demonstrating what the previous designation of this wetland was
#  - Wetland DEM: Environmental layer showing what the measured DEM is at Trumbull Creek Wetland
#  - Phosphorus Data: Measured vector data from an undergrad researcher in my lab (Emily)


# One of the undergrads in my lab group collected data at the same wetland I did, I will be using some of her SURE work and data points for this exercise
emily_data <- read_csv("./lab5_data/EmilyCampbell_Combined_SURE_Work.csv")

#turning her work in an sf object
emily_data_sf <- st_as_sf(emily_data, coords = c("longitude", "latitude"), crs = 4326)

#NE Ohio national landuse cover data
neoh_nlcd <- terra::rast("./lab5_data/neoh_nlcd.tif") 

#DEM of the wetland area (TRUC aka Trumbull Creek) itself
TRUC_dem <- terra::rast("./lab5_data/dem_partially_filled.tif")

# Study area of project - TRUC parcel
boundary <- sf::read_sf("./lab5_data/PROJECT_LIMITS.shp") 

#checking crs of these rasters and shapefile
st_crs(neoh_nlcd) #Albers Conic
st_crs(TRUC_dem) #WGS 84, zone 17N
st_crs(boundary) #NAD 83


#reprojecting boundary to WGS 84
boundary_new <- st_transform(boundary, crs= 4326)

TRUC_dem_new <- TRUC_dem %>%  #take raster in 
  terra::project(., boundary_new) %>% #project it to the same as study area
  terra::crop(., boundary_new) %>% #crop it to boundary bbox
  terra::mask(., boundary_new)  #mask to actual boundaries

neoh_nlcd_new <- neoh_nlcd %>%  #take raster in 
  terra::project(., boundary_new, method = "near") %>% #project it to the same as sa
  #Had to do this because it is resampling and has decimal values for nlcd
  terra::crop(., boundary_new) %>% #crop it to boundary bbox
  terra::mask(., boundary_new) #mask to actual boundaries

# cropping emily's points to only within this parcel boundary
emily_points_inside <- st_filter(emily_data_sf, boundary_new)


```

### 3.2 — Build the interactive map

Build the Leaflet map with the following requirements:

- At least **two distinct data layers** (one social, one environmental)
- A **layer toggle control** (`addLayersControl()` with `overlayGroups`) that lets the user turn each layer on or off independently
- At least **one additional interactive feature** of your choosing that is meaningful for YOUR specific map. Options include (but are not limited to):
  - Popups with attribute data
  - Hover labels
  - A minimap (`addMiniMap()` from `leaflet.extras`)
  - Marker clustering
  - A scale bar (`addScaleBar()`)
  - Highlighting on hover (`highlight` argument in `addPolygons()`)
- A legend for at least one layer

Document your "one thing" choice in a comment explaining what it is and why it fits your map.

```{r task3_map, echo=TRUE, message=TRUE}
# My "one thing": I added attribute data pop ups for this map. I wanted the size of the yellow circles to indicate how much phosphorus (mg/kg soil) was measured at each location, but it can vary widely throughout the site. Having the pop ups helps keep that data associated with the map without clustering too much. I don't think a legend would have conveyed that data very well.

# I need to convert all of my raster files using the Raster package --> not terra
dem_raster <- raster::raster(TRUC_dem_new)  # convert terra -> raster
nlcd_raster <- raster::raster(neoh_nlcd_new)

# Creating color palettes for each of my rasters
# DEM: using a continuous terrain palette to visualize it
pal_dem <- colorNumeric(palette = terrain.colors(7), 
                        values(dem_raster), 
                        na.color = "transparent")

# NLCD: categorical, I'm choosing to map NLCD codes to standard colors
nlcd_vals <- unique(values(nlcd_raster)) #makes sure it doesn't repeat over again
pal_nlcd <- colorFactor(palette = "Set3", 
                        domain = nlcd_vals, 
                        na.color = "transparent")
#popup of the phosphorus data
emily_points_inside <- emily_points_inside %>%
  mutate(
    popup_lisa_new = paste0(
      "H2Ohio Sampling ID: <b>", H2Ohio_ID, " </b><br>",
      "Phosphorus (mg/kg soil): <b>", round(P_mg_kgsoil, 2), "</b><br>"))

# Creating the actual map now
leaflet() %>% #leaving this blank since its not one data set
    addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>% #basemap 1
    addProviderTiles(providers$OpenStreetMap, group = "OSM") %>% #basemap 2
  addRasterImage(dem_raster, # raster input
                 colors = pal_dem, 
                 opacity = 0.7, 
                 group = "DEM of TRUC") %>%
  addRasterImage(nlcd_raster, #NLCD raster!
                 colors = pal_nlcd, 
                 opacity = 0.7, 
                 group = "NLCD Classifications") %>%
   addPolygons(data = boundary_new, #boundary parcel of TRUC 
              color = "blue", 
              weight = 4, 
              fill = FALSE, 
              group = "Study Area") %>%
  addCircleMarkers(data = emily_points_inside,
                   radius = ~P_mg_kgsoil, #base size off of amount of Phosphorus
                   color = "yellow", 
                   fillOpacity = 0.9,
                   popup       = ~popup_lisa_new,  
                   group = "Sample Points") %>%
  addLegend(pal = pal_dem, #DEM legend and palette
            values = values(dem_raster),
            title = "Elevation (m)", 
            position = "bottomright", 
            group = "DEM of TRUC") %>%
  addLegend(pal = pal_nlcd,
            values = nlcd_vals,
            title = "NLCD Land Cover", 
            position = "bottomleft", 
            group = "NLCD Classifications") %>%
  addLayersControl(
    baseGroups = c("Satellite", "OSM"),
    overlayGroups = c("DEM of TRUC", "NLCD Classifications", "Study Area", "Sample Points"),
    options = layersControlOptions(collapsed = FALSE))



```
**color brewer palettes https://r-graph-gallery.com/38-rcolorbrewers-palettes.html
https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/



**Reflection question 3 (answer in your repo):** Describe the one extra feature you added and justify why it is meaningful for your specific map — not just technically interesting, but cartographically or analytically useful for the story you are trying to tell.
- The extra feature I added was not that complex, it was just pop ups that describe the amount of phosphorus measured and the H2Ohio sampling ID that was used. Having this popup would be really helpful for people in my lab group that would want to take a quick glance at a map and where the zones of phosphorus might be clustered in relation to the elevation or proximity to water source. I think having this information displayed in a table or legend can sometimes make it less visually appealing and quite busy, especially if there are many other variables or data sources being shown. I think having these as popups keep its a bit less busy, but allows the user to look more in depth if they choose. 

---

## Questions

1. Reflect on the labs from this semester. What did you learn? What did you like? What did you not like? (Be specific — vague answers will receive partial credit.)
- I feel like I learned a lot throughout the year. At the start I had a very vague, but broad understanding of geospatial analysis and coding from a project I did in my undergrad. I learned many new practices and techniques like GitHub, organizing my data, dynamic vs static mapping, and more about the tidyverse world. I really enjoyed getting to practice some of these skills using real data from either Dr. Bitterman or my own research because I can use it to help interpret new information I collect later on in my dissertation.
-I also spent very little time making functions and understanding how they worked before this class so that was really huge for me to be challenged to write my own and now seeing areas in my old code where I could have used them has been super eye opening. - I didn't love spending so much time in tmap, I completely understand why we are taught that and it can be a helpful tool, I wish we could have spent more time working in ggplot and seeing some of the ggspatial code, but with the the understanding I have now, I could easily apply it.

2. In this lab, all three tasks involved taking existing analysis and presenting it interactively. What does Leaflet add beyond what `tmap` could do? And conversely, what did you give up by switching to Leaflet? Consider both technical and cartographic dimensions.
- Leaflet enables us advanced customization with specific palettes, fills, colors, labeling schemes, and the ability to have minimaps, popups, highlighting, and other visually significant cartographic features. It also allows people the ability to toggle through several different created maps without needing several separate printouts or overly busy static and somewhat dynamic tmaps. It allows people to sift through different data with a simple click and to view the world along side of it, instead of just a plain background or a crazy view port. 
- Leaflet has many different plugins, many of which I haven't tackled yet, like the swiping feature and drawing and editing tools. This takes data visualization beyond the classic bounds of tmap. However, leaflet can be quite picky and often doesn't play well with raster data and needs to use the outdated Raster package or changed visualization to have the same flexibility that tmap has built into it. 
- In the same view, tmap is quite simple, easily layered, and has similar syntax to ggplot which allows more people to quickly access it to visualize data. Tmap can become interactive which gives it the edge of both static and dynamic mapping where leaflet is only used for dynamic mapping. Along with that, for people who don't need a lot of customization, tmap quickly handles formatting and palettes with the styles functionality.


3. Describe the "one thing" you chose to add to your map in Task 3. What did you implement, why is it applicable to your specific map, and how does it change the user's experience compared to a static version?
-The extra feature I added was not that complex, it was just pop ups that describe the amount of phosphorus measured and the H2Ohio sampling ID that was used. Having this popup would be really helpful for people in my lab group that would want to take a quick glance at a map and where the zones of phosphorus might be clustered in relation to the elevation or proximity to water source. 
-Having this information traditionally displayed in a table or legend can make it far less visually interesting and harder to interpret without having the relative size of the circle markers corresponding to the amount of phosphorus measured and having the pop up there to only appear when needed. The static version would have had to had the circles labeled which is often quite busy and hard to read, especially at a site this small with samples taken so closely together, but it also limits what you can display. i chose to just show the H2Ohio ID along with the Phosphorus, but those pop ups could have been much longer and customized with the metals count, the year it was sampled, a whole suite of environmental details, etc. Having this as a dynamic map allows the user to have a more interactive experience and have much more control over what portions of this data (ecological and sociological) that they are viewing and interpreting at a single point in time. It also is handy for when people might have questions about the terrain in relation to either of the basemaps (proximity to features identified through satellite or cities).

---

## Bonus (+20 points)

Extend your Task 2 LISA map to include a **synchronized side-by-side view**: the LISA quadrant map on one side and the underlying variable choropleth (from Task 1 or your Lab 3 variable) on the other. The `leafsync` package provides `sync()` for this purpose.

The two maps should share the same extent and zoom level. When you pan or zoom one, the other follows. Add a brief comment explaining what analytical value this comparison provides.

```{r bonus, echo=TRUE, message=TRUE}
# your code here
```
