---
title: "Lab 06"
#author: "Dr. Patrick Bitterman"
#date: "`r format(Sys.time(), '%Y %B %d')`"
output: html_document
my_subtitle: "Env Analysis in R"
geometry: margin=1in
---

# Lab 06: Dynamic mapping with Leaflet

### Read the instructions COMPLETELY before starting the lab

This lab builds directly on labs 2–4 and the in-class Leaflet exercises. You will need those lab files available for reference.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

*Note: This lab is output as an HTML document — not a PDF. Leaflet maps are **interactive** and cannot be embedded in a static PDF. Your knitted output will be an `.html` file. Keep this in mind when organizing your submission.*

## Introduction

In the previous lab you built static maps using `tmap`. Static maps are excellent for print, publication, and controlled presentation. But many modern mapping applications demand **interactivity**: users want to pan, zoom, click, and toggle layers. Leaflet is the dominant open-source library for interactive web maps, and `leaflet` (the R package) wraps it for use in R.

Leaflet's design philosophy is similar to `tmap` in that maps are built by **layering components**:

```
leaflet() %>%
  addTiles() %>%                  # base map (tiles)
  addPolygons(...) %>%            # vector layer
  addLegend(...) %>%              # legend
  addLayersControl(...)           # layer toggle
```

Each `add*()` call adds a new layer or control to the map object. The `%>%` pipe passes the map forward. This means you can build up a map incrementally, verify each piece, and compose them together — exactly as you did in `tmap`.

Key resources:
- Official leaflet for R docs: https://rstudio.github.io/leaflet/
- Making maps with R (Hahn, Ch. 5): https://bookdown.org/nicohahn/making_maps_with_r5/docs/leaflet.html
- `leaflet.providers` for basemap tile options

```{r packages, echo=TRUE, message=TRUE}
library(sf)
library(leaflet)
library(tidyverse)
library(spdep)   # needed for Task 2 (Moran/LISA)
```

*Note: You may need additional packages depending on your Task 3 choices. Load them here.*

---

## Task 1: Choropleth from Lab 2

*Reference: leaflet docs — `addPolygons()`, `colorBin()`, `addLegend()`*

In Lab 2, Task 2.3, you created a choropleth of Chesapeake Bay Watershed counties shaded by the total cost of BMPs funded/implemented in each county. Here you will recreate that map in Leaflet.

### 1.1 — Reload and prepare the data

Go back to your Lab 2 code and reload the county geometry and BMP cost data. Perform whatever join and preprocessing was necessary to produce your choropleth variable. Store the result as an `sf` object called `counties_bmp`.

Before mapping, **check and set the CRS**. Leaflet requires geographic coordinates (WGS 84, EPSG 4326). If your data are in a projected CRS, reproject now using `sf::st_transform(counties_bmp, 4326)`.

```{r task1_data, echo=TRUE, message=TRUE}
# Load in the county data set from CBW
counties <- sf::read_sf("./CBW/County_Boundaries.shp") %>% 
  sf::st_make_valid()

#bmp cost data
bmps <- read_csv("./CBW/BMPreport2016_landbmps.csv")

 #getting the county FIPS code pulled out of BMPs 
unique_bmps <- bmps |> mutate(COUNTYFP10 = stringr::str_sub(GeographyName, 3, 5), # matching columns
                       STATEFP10 = stringr::str_sub(GeographyName, 1, 2)) |>
                       group_by(STATEFP10, COUNTYFP10) |>  #need to group by both
                summarize(TotalCost = sum(Cost, na.rm=TRUE), .groups="drop") |> #total cost per county
                distinct() #keep only the unique state and counties

#unique counties
unique_counties <- counties |>
  select(STATEFP10, COUNTYFP10, geometry, NAMELSAD10) |> 
  distinct() #keeping only the unique counties (not all entries)

#combine counties and BMPS by STATE and COUNTY FP10
counties_bmp <- unique_counties |> left_join(unique_bmps, by = c("COUNTYFP10","STATEFP10")) 

#check crs of the counties_bmp
sf::st_crs(counties_bmp) #looks all good, its in WGS 84

```

### 1.2 — Define a color palette

Leaflet's `colorBin()` function creates a color palette from a set of breaks. You must determine the **equal-interval breaks programmatically** — do not hard-code them.

*Hint: The range of the variable divided by the number of classes gives you the interval width. `seq()` can generate your break points.*

```{r task1_palette, echo=TRUE, message=TRUE}
#define the breaks
breaks <- seq(min(0), max(200000000), length.out=7) # break points, 7 seemed best
# has it broken down to this range representing millions

# palette
pal <- colorBin(palette = "BuPu", domain = 0:200000000, bins = breaks) #blue to purple palette
```

*Requirement: Do NOT use the default color scheme (Blues). Choose a palette appropriate for sequential data from a source like ColorBrewer.*

### 1.3 — Build the choropleth map

Build the Leaflet map with the following requirements:

- The county polygons colored by total BMP cost using your `colorBin()` palette
- A **mouse-over label** that displays the county name and total cost when the cursor hovers over a county. Use the `label` argument in `addPolygons()`.
- A **legend** using `addLegend()`
- A basemap tile layer using `addProviderTiles()` — choose something other than the default OpenStreetMap. Browse options at https://leaflet-extras.github.io/leaflet-providers/preview/

```{r task1_map, echo=TRUE, message=TRUE}
#Map it outttt
leaflet(counties_bmp) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap,
                   group = "NatGeo World Map")  %>% #base map 1
  addProviderTiles(providers$Esri.WorldImagery,
                   group = "Satellite View")  %>% #base map 2
  addPolygons( #start with choropleth map
    group = "BMP Map",
      fillColor = ~pal(TotalCost), #my pal and total cost
      fillOpacity = 0.9,
      color = "black",    
      weight = 1,
      label = ~paste0(NAMELSAD10, ": $", #prints the county name with a space after
                      formatC(TotalCost, format = "f", digits = 0, big.mark = ","))) %>%
                      #has the total cost after, no decimals, just commas
  addLegend( #legend for the  choropleth
  position = "bottomright",
  title    = "BMPs Funded/ Implemented in Each County",
  pal   = pal,
  values   = ~TotalCost,
  opacity  = 1,
  labFormat = labelFormat(prefix = "$", big.mark = ",")) %>% #fixes scientific notation
  addLayersControl(
    baseGroups = c("NatGeo World Map", "Satellite View"), 
    overlayGroups = ("BMP Map"), 
    options = layersControlOptions(collapsed = FALSE))

```

**Reflection question 1 (answer in your repo):** In `tmap`, classification method choices (jenks, quantile, etc.) are made when you call `tm_polygons()`. In Leaflet, you define the palette and breaks separately using `colorBin()`, `colorQuantile()`, or `colorNumeric()`. What are the trade-offs of this approach? Does separating the palette from the geometry layer give you more or less flexibility? Give a specific example.

- I think there are some tradeoffs by using these two methods, but in general I think separating the palette from the geometry layer could give you more flexibility. In scenarios when you need to have lots of control over the palettes or ranges, using the leaflet method allows you to preset different colors and breaks to the data which can be resued over time in different maps. This allows a consistent and steady color scheme and helps with publishable figures. Also, being able to control it for multiple different layers can help you keep your code tidy and functioning. On the other hand, it is very simple and quick to just toss in the method choice within the tmap settings, which allows for quick, easy plotting if the map is just a check for future work.

---

## Task 2: LISA map from Lab 3

*Reference: leaflet docs — `addPolygons()` popup, `addLayersControl()`, `addProviderTiles()`*

In Lab 3, Bonus Task 2, you produced a choropleth map of LISA (local Moran's I) quadrants — a categorical four-class map showing spatial clustering patterns (HH, HL, LH, LL). Here you will recreate that map in Leaflet with interactive enhancements.

### 2.1 — Reload and prepare the data

Go back to your Lab 3 code. Reload the spatial data and rerun the Moran/LISA analysis to recover the quadrant assignments and p-values. Store the result as an `sf` object called `counties_lisa`, reprojected to WGS 84 (EPSG 4326).

*Note: You need both the quadrant labels and the p-values — make sure both are columns in `counties_lisa` before proceeding.*

```{r task2_data, echo=TRUE, message=TRUE}
#choosing to look at female ages 22-24 population per county in those CO, UT, NM, KS, WY 
#reading in the exported table from that task
f.pop.states.1 <- sf::read_sf("./f.pop.states.gpkg") 

f.pop.states.1 <- f.pop.states.1 |>
  mutate(female_proportion = (B01001_034E/B01001_001E)) # female (22-24) proportion of total pop in each county

# Build spatial weights matrix
coords <- sf::st_geometry(f.pop.states.1)
nb  <- spdep::poly2nb(f.pop.states.1, queen = TRUE)

#row standardize the weights
lw  <- spdep::nb2listw(nb, style = "W", zero.policy = TRUE)

# Compute Local Moran's I
lisa <- spdep::localmoran(f.pop.states.1$female_proportion, lw, zero.policy = TRUE)
f.pop.states.1$local_I  <- lisa[, "Ii"] #moran's i column
f.pop.states.1$lisa_pval <- lisa[, "Pr(z != E(Ii))"] #moran's pvalue

#LISA Quadrant Labels
# Mean-center the variable and the spatial lag
x     <- f.pop.states.1$female_proportion
x_lag <- lag.listw(lw, x, zero.policy = TRUE)
x_mean     <- mean(x,     na.rm = TRUE)
xlag_mean  <- mean(x_lag, na.rm = TRUE)


# create the quadrants
counties_lisa <- f.pop.states.1 %>%
  mutate(
    quad = case_when(
      x >  x_mean & x_lag >  xlag_mean ~ "HH",
      x >  x_mean & x_lag <= xlag_mean ~ "HL",
      x <= x_mean & x_lag >  xlag_mean ~ "LH",
      x <= x_mean & x_lag <= xlag_mean ~ "LL"))


```

### 2.2 — Define a categorical color palette

LISA quadrants are a categorical variable (four classes). Use `colorFactor()` instead of `colorBin()`. Assign a distinct, meaningful color to each quadrant (HH, HL, LH, LL). A common convention is warm colors for High clusters and cool colors for Low clusters.

```{r task2_palette, echo=TRUE, message=TRUE}
# making palette for the lisa, not coding it into the df
pal_lisa <- colorFactor(palette = c("#d7191c","#fdae61","#abd9e9","#2c7bb6"), 
                        levels = c("HH", "HL", "LH", "LL"))
```

### 2.3 — Build the LISA map

Build the Leaflet map with the following requirements:

- County polygons colored by LISA quadrant using your `colorFactor()` palette
- A **pop-up window** (not a hover label) that displays the p-value for that county when clicked. Use the `popup` argument in `addPolygons()`. Format the p-value to a reasonable number of decimal places.
- A **basemap switcher** that allows the user to toggle between **3 different basemaps**. Use `addProviderTiles()` with `group` arguments and `addLayersControl()` with `baseGroups`.
- A legend

```{r task2_map, echo=TRUE, message=TRUE}

#Popup to get the moran's i and the p-value
counties_lisa <- counties_lisa %>%
  mutate(
    popup_lisa = paste0(
      "<b>", NAME, " </b><br>",
      "LISA Quadrant: <b>", quad, "</b><br>",
      "Local Moran p-value: <b>", round(lisa_pval, 4), "</b>" ))

#check the crs before plotting
st_crs(counties_lisa) # its the 1983 datum so it needs reprojected for leaflet

#transform to leaflet crs
counties_lisa <- st_transform(counties_lisa, crs = 4326)

#turn quad into factor 
counties_lisa$quad <- as.factor(counties_lisa$quad)

#Map
leaflet(counties_lisa) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap,
                   group = "NatGeo World Map")  %>% #base map 1
  addProviderTiles(providers$Esri.WorldImagery,
                   group = "Satellite View")  %>% #base map 2
  addProviderTiles(providers$Stadia.StamenWatercolor,
                   group = "Stamen Watercolor")  %>% #base map 3
  addPolygons( #LISA quadrant map
    group= "LISA Quadrant",
    fillColor   = ~pal_lisa(quad),
    fillOpacity = 0.9,
    color       = "black",
    weight      = 2,
    popup       = ~popup_lisa,
    highlightOptions = highlightOptions(
      weight       = 2,
      fillOpacity  = 0.9,
      bringToFront = TRUE
    )) %>%
  addLegend(
  position = "bottomright", #stacking legends
  title    = "LISA Quadrant",
  colors   = pal_lisa(unique(counties_lisa$quad)), #this makes it just appear once
  labels   = unique(counties_lisa$quad), #this makes so that it only prints the values once, not 5 times
  opacity  = 0.9) %>%
  addLayersControl(
    baseGroups = c("NatGeo World Map", "Satellite View", "Stamen Watercolor"), 
    overlayGroups = "LISA Quadrant", 
    options = layersControlOptions(collapsed = FALSE))
```

*Tip: The `addLayersControl()` function accepts `baseGroups` (radio buttons, one active at a time) and `overlayGroups` (checkboxes, multiple can be active). Basemaps should be `baseGroups`.*

**Reflection question 2 (answer in your repo):** Compare the static LISA map you made in Lab 3 to this interactive version. What does interactivity add? Are there things the static map communicates more clearly? Consider map reading purpose (exploration vs. communication) in your answer.
-I feel like this interactive map is better for the exploration of this data and how these five states are related to each other. Unless this was published to a website, it would be a lot more difficult to get people to see this map compared to a static map that could be published in a manuscript or something else. The fact that this leaflet map has popups makes it far more convenient for showing different spatial patterns without overflowing the map with county names, p-values, and what the clusters are. It is also helpful to toggle between different basemaps to see if other factors are influencing this behavior like placement of mountain ranges, distance to the coastline, surrounding states, etc.

---

## Task 3: Free-form map from Lab 5

*Reference: leaflet docs — `addLayersControl()` with overlayGroups, your choice of other features*

In Lab 5, Task 5, you created an original map combining vector and raster data, representing both a social and an environmental process. Here you will recreate that concept in Leaflet.

*Note: Leaflet handles raster data differently than `tmap`. The `addRasterImage()` function in the `leaflet` package can display projected rasters, but requires the `raster` package (not `terra`) or conversion. An alternative is to represent the "environmental process" through a different vector layer (e.g., watersheds, land cover categories as polygons) if raster integration proves difficult. Document your approach in a comment.*

### 3.1 — Reload and prepare your data

Reload the data you used in your original free-form map. Reproject all layers to WGS 84 (EPSG 4326). List the layers you are using in a comment.

```{r task3_data, echo=TRUE, message=TRUE}
# To remake the map from lab 5 I will be using some data from my wetland lab group
# layers I'm using:
#  - Landuse Landcover: Social layer demonstrating what the previous designation of this wetland was
#  - Wetland DEM: Environmental layer showing what the measured DEM is at Trumbull Creek Wetland
#  - Phosphorus Data: Measured vector data from an undergrad researcher in my lab (Emily)


# One of the undergrads in my lab group collected data at the same wetland I did, I will be using some of her SURE work and data points for this exercise
emily_data <- read_csv("./lab5_data/EmilyCampbell_Combined_SURE_Work.csv")

#turning her work in an sf object
emily_data_sf <- st_as_sf(emily_data, coords = c("longitude", "latitude"), crs = 4326)

#NE Ohio national landuse cover data
neoh_nlcd <- terra::rast("./lab5_data/neoh_nlcd.tif") 

#DEM of the wetland area (TRUC aka Trumbull Creek) itself
TRUC_dem <- terra::rast("./lab5_data/dem_partially_filled.tif")

# Study area of project
boundary <- sf::read_sf("./lab5_data/PROJECT_LIMITS.shp") 

#checking crs of these rasters
st_crs(neoh_nlcd)
st_crs(TRUC_dem)

#reprojecting boundary to the same as SA from earlier
boundary_new <- st_transform(boundary, st_crs(sa))

TRUC_dem_new <- TRUC_dem %>%  #take raster in 
  terra::project(., boundary_new) %>% #project it to the same as sa
  terra::crop(., boundary_new) %>% #crop it to boundary bbox
  terra::mask(., boundary_new)  #mask to actual boundaries

neoh_nlcd_new <- neoh_nlcd %>%  #take raster in 
  terra::project(., boundary_new, method = "near") %>% #project it to the same as sa
  #Had to do this because it is resampling and has decimal values for nlcd
  terra::crop(., boundary_new) %>% #crop it to boundary bbox
  terra::mask(., boundary_new) #mask to actual boundaries

#getting the real nlcd color palette
nlcd_cols <- c(
  "11" = "#476BA1",  # Open Water
  "21" = "#DDC9C9",  # Developed, Open Space
  "22" = "#D89382",  # Developed, Low Intensity
  "23" = "#ED0000",  # Developed, Medium Intensity
  "24" = "#AA0000",  # Developed, High Intensity
  "31" = "#B2ADA3",  # Barren
  "41" = "#68AA63",  # Deciduous Forest
  "42" = "#1C6330",  # Evergreen Forest
  "43" = "#B5C98E",  # Mixed Forest
  "52" = "#CCBA7C",  # Shrub/Scrub
  "71" = "#E2E2C1",  # Grassland/Herbaceous
  "81" = "#DED83E",  # Pasture/Hay
  "82" = "#AA7028",  # Cultivated Crops
  "90" = "#BAD8EA",  # Woody Wetlands
  "95" = "#70A3BA")   # Emergent Herbaceous Wetlands

#had issues with the categorical raster, now I am converting it to a factor first
# Convert NLCD raster to categorical
neoh_nlcd_new <- terra::as.factor(neoh_nlcd_new)


#creating a map showing the dem of TRUC (environmental), the landuse designation (social), and emily's data points(vector)

TRUC_map <-
  tm_shape(TRUC_dem_new) + #raster dem of TRUC
    tm_raster(col.scale = tm_scale_continuous(values = "terrain"),
              col.legend = tm_legend(position = tm_pos_out("center", "bottom"),
                                     title="Elevation",
                                     orientation="landscape")) + #legend on the bottom
  tm_shape(boundary_new) +  #TRUC shapefile parcel boundary
    tm_fill(fill_alpha = 0.2) +
    tm_borders(col = "black") +
  tm_shape(neoh_nlcd_new) +
    tm_raster( #for some reason the colors are breaking and are not plotting on here :(
          col.scale = tm_scale_categorical(values = nlcd_cols), #categorical data and my color palette
          col_alpha = 0.6, #slightly transparent
          title = "Historic NLCD Land Cover Classes") +
      tm_scalebar(position = c("center", "bottom")) +
      tm_compass(position = c("right", "top"), size = 1.5) +
  tm_shape(emily_data_sf) +
    tm_dots(size="P_mg_kgsoil", fill="gold",
            size.legend = tm_legend("Measured Phosphorus per mg/kg of Soil")) +
  tm_layout(#I can't seem to get this title any bigger, tried both tm_title and tm_layout
      main.title = "Soil Phosphorus Measured Throughout the Newly Restored
Trumbull Creek Wetland and its Previous National Land Cover Class Designation",
      main.title.size =0.7, #any higher shoots the title away from the map
      main.title.position = "center",
      main.title.fontface = "bold")

TRUC_map 

```

### 3.2 — Build the interactive map

Build the Leaflet map with the following requirements:

- At least **two distinct data layers** (one social, one environmental)
- A **layer toggle control** (`addLayersControl()` with `overlayGroups`) that lets the user turn each layer on or off independently
- At least **one additional interactive feature** of your choosing that is meaningful for YOUR specific map. Options include (but are not limited to):
  - Popups with attribute data
  - Hover labels
  - A minimap (`addMiniMap()` from `leaflet.extras`)
  - Marker clustering
  - A scale bar (`addScaleBar()`)
  - Highlighting on hover (`highlight` argument in `addPolygons()`)
- A legend for at least one layer

Document your "one thing" choice in a comment explaining what it is and why it fits your map.

```{r task3_map, echo=TRUE, message=TRUE}
# your code here

# My "one thing": [describe what you added and why it's applicable]
```

**Reflection question 3 (answer in your repo):** Describe the one extra feature you added and justify why it is meaningful for your specific map — not just technically interesting, but cartographically or analytically useful for the story you are trying to tell.

---

## Questions

1. Reflect on the labs from this semester. What did you learn? What did you like? What did you not like? (Be specific — vague answers will receive partial credit.)

2. In this lab, all three tasks involved taking existing analysis and presenting it interactively. What does Leaflet add beyond what `tmap` could do? And conversely, what did you give up by switching to Leaflet? Consider both technical and cartographic dimensions.

3. Describe the "one thing" you chose to add to your map in Task 3. What did you implement, why is it applicable to your specific map, and how does it change the user's experience compared to a static version?

---

## Bonus (+20 points)

Extend your Task 2 LISA map to include a **synchronized side-by-side view**: the LISA quadrant map on one side and the underlying variable choropleth (from Task 1 or your Lab 3 variable) on the other. The `leafsync` package provides `sync()` for this purpose.

The two maps should share the same extent and zoom level. When you pan or zoom one, the other follows. Add a brief comment explaining what analytical value this comparison provides.

```{r bonus, echo=TRUE, message=TRUE}
# your code here
```
