---
title: "Lab 04"
#author: "Dr. Patrick Bitterman"
#date: "`r format(Sys.time(), '%Y %B %d')`"
output: pdf_document
my_subtitle: "Env Analysis in R"
fancy: true
geometry: margin=1in
latex_engine: pdflatex
colorlinks: true
---

# Lab 04: Raster data management and analysis

### Read the instructions COMPLETELY before starting the lab

This lab builds directly on the raster analysis concepts we worked through in class, including the Week 12 in-class notebook and the Wildcard Thursday activity. You should have that code available for reference.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

## Data

All data for this lab are in the course repository at:

../data/wildcard_thursday/erie_raster_ex/


You will find two subdirectories:

- `images/` — seven weekly composite GeoTIFF rasters of cyanobacteria bloom intensity in Lake Erie (summer–fall 2017)
- `bounds/` — several shapefiles representing geographic areas of interest within the Lake Erie region

**Before writing any code, read the metadata file at**

**`../data/wildcard_thursday/erie_raster_ex/erie_README.txt`.** 

This is not optional. Understanding your data source, its scaling, and its data key is essential for correct analysis.

## Introduction

Cyanobacteria (blue-green algae) blooms in Lake Erie have become a significant environmental and public health concern. These blooms can produce toxins that contaminate drinking water and harm aquatic ecosystems. The data you will use were produced by NOAA using imagery from the European Space Agency's Sentinel-3 satellite. Each raster represents the maximum observed bloom intensity during a given week.

The product you will be working with is the **CIcyano** (Cyanobacteria Index, cyano only) product. Values in these rasters are stored as **digital numbers (DN)** and must be transformed to meaningful physical units before interpretation. You saw this in class — recall the reverse-scaling formula from the README and from the Week 12 notebook.

A critical aspect of working with remote sensing data is understanding **sentinel values**: special numeric codes that indicate conditions like cloud cover, land pixels, or missing data. These are NOT real observations and must be handled carefully before any analysis.

We begin by loading the relevant packages:

```{r packages, echo=TRUE, message=TRUE}
library(terra)
library(sf)
library(tidyverse)
library(tmap)
library(fs)
```

---

## Task 1: Data management and preprocessing

### 1.1 Inventory your rasters

Use `fs::dir_ls` with the parameter `regexp = "\\.tif$"` to get the file paths for all raster images in the `images/` directory. Store the result in an object called `raster_files`. Print the result — how many files are there?

```{r list_files, echo=TRUE, message=TRUE}
#Finding files with the .tif extension
raster_files <- fs::dir_ls("erie_raster_ex/images/", regexp = "\\.tif$")
print(raster_files) #tehre are 7 tif images
```

### 1.2 Inspect a single raster

Load the first raster in your list using `terra::rast()`. Examine the following properties and record what you find:

- Coordinate reference system: `terra::crs()`
- Spatial extent: `terra::ext()`
- Spatial resolution: `terra::res()`
- Number of layers: `terra::nlyr()`
- Range of cell values (use `terra::global()` with `"min"` and `"max"`)

```{r inspect_raster, echo=TRUE, message=TRUE}
#load in first raster from list
rast_1 <- terra::rast("erie_raster_ex/images/ts_2017.0802_0808.L4.LE3.CIcyano.MAXIMUM_7day.tif")

#check crs
terra::crs(rast_1)

#spatial extent
terra::ext(rast_1)

#spatial resolution
terra::res(rast_1)

#number of layers in raster 
terra::nlyr(rast_1)

#range of cell values
terra::global(rast_1, c("min", "max")) #got only NA values
terra::global(rast_1, c("min", "max"), na.rm=TRUE)

```
Answers:
-CRS= WGS 84, UTM zone 17N
-Extent= 281282.934996, 694382.934996, 4572608.997997, 4764308.997997 (xmin, xmax, ymin, ymax)
-Resolution = 300 x 300 meters
-Layers= 1
-Range of cell values (minus NA) = 1-252



### 1.3 Write a preprocessing function: `clean_cicyano()`

Looking at the data key in the README, you will notice that values of 0 and 250 and above have special meanings — they are **not** valid bloom observations. Before doing any analysis, these sentinel values must be replaced with `NA`.

Write a function called `clean_cicyano(r)` that:

1. Takes a SpatRaster `r` as input
2. Builds a reclassification matrix that maps **0** and **all values >= 250** to `NA`, and leaves values 1–249 unchanged
3. Applies the reclassification using `terra::classify()`
4. Returns the cleaned raster

*Hint: Your reclassification matrix should handle the no-detection case (0) and all sentinel codes (250 through 255). Review how `terra::classify()` uses a three-column matrix to specify `[from, to, becomes]` intervals. You may find `right = NA` useful here so boundary values like 0 and 250 are included in the intervals.*


```{r clean_function, echo=TRUE, message=TRUE}
clean_cicyano <- function(r) {
  
  #make reclassification matrix - only keeping values I want
  rcl_matrix <- matrix(c(0,0,NA,
                         250,255,NA),ncol = 3, byrow = TRUE)
  
  #cleaned  up raster, setting invalid data to NA
  cleaned_raster <- terra::classify(r, rcl = rcl_matrix, right=NA)
  
  return(cleaned_raster)
}

#test function
test1<-clean_cicyano(rast_1)
plot(test1)

terra::freq(test1)
#comparing min
terra::global(test1,"min", na.rm=TRUE)
terra::global(rast_1,"min", na.rm=TRUE)
#comparing max
terra::global(test1,"max", na.rm=TRUE)
terra::global(rast_1,"max", na.rm=TRUE)

```

Test your function on the first raster. Verify it worked by comparing the global min/max of the raw raster vs. the cleaned raster.

### 1.4 Write a transform function: `dn_to_ci()`

The DN values in these rasters are scaled according to the formula in the README. To convert back to physical CI units, use the reverse-scaling formula:

$$CI = 10^{\left(\frac{3.0}{250.0} \times DN - 4.2\right)}$$

Write a function called `dn_to_ci(r)` that takes a cleaned SpatRaster and returns a new raster with values transformed to CI units. You can apply a function to a raster using standard arithmetic operations or by passing a function to the raster.

```{r transform_function, echo=TRUE, message=TRUE}
dn_to_ci <- function(x) {
   10**(((3.0 / 250.0) * x) - 4.2)
}

#passing a function to the raster
test.ci <- test1 %>% dn_to_ci()
plot(test.ci)

#checking the range
terra::global(test.ci,c("min","max"), na.rm=TRUE)

```

Test your function. What is the range of CI values in the transformed raster? Do they make physical sense given the data key?
-The range appears to be a very small number close to 0 up to 0.061. This range does make sense because it is showing high bloom intensity in areas that are normally breaking out in the western basin of Lake Erie and Sandusky Bay. It has most of the lake with no bloom severity which makes sense contextually.

### Questions for Task 1:

**Q1.1** What does the CRS of these rasters tell you about how distances and areas are measured? Would you need to reproject before calculating bloom area in square kilometers?
-The crs of these rasters tells us that area and distance is preserved pretty well within zones, but some distortion could be experienced when measuring across zones. It is already in meters so you can calculate the bloom area without needing to reproject.

**Q1.2** Why is it important to remove sentinel values *before* applying the DN-to-CI transform — rather than after? What would happen to a land pixel (value = 252) if you transformed it without cleaning first?
-It is important because those values could be incorrectly counted as bloom area, meaning that land cover is now part of the bloom radius. That pixel would be transformed into a very high level bloom around 0.067.

**Q1.3** Why is the `right = NA` argument useful in `terra::classify()` for this task?
- This function essentially closes the intervals on both sides in a from-to-becomes classification like in the question.This allows 1 and 249 to be included in the values that are unchanged without having to write out 0-1, 1-249,249-255, etc.
---

## Task 2: Spatial subsetting — crop and mask

### 2.1 Load and align an area of interest

Load the `bb_erie_sandusky` shapefile from the `bounds/` directory using `sf::read_sf()`. Check its CRS. If it does not match the raster CRS, reproject it using `sf::st_transform()`.

```{r load_aoi, echo=TRUE, message=TRUE}
#read in shp
erie_sand <- sf::read_sf("erie_raster_ex/bounds/bb_erie_islands.shp")
#check crs
terra::crs(erie_sand) #it has the same crs as the raster
terra::crs(rast_1)
```

### 2.2 Crop and mask

Using the **August 16–22, 2017** raster (`ts_2017.0816_0822`):

1. Apply `clean_cicyano()` and `dn_to_ci()` to get a clean, transformed raster
2. Use `terra::crop()` to clip the raster extent to the Sandusky Bay boundary
3. Use `terra::mask()` to set pixels *outside* the boundary to `NA`

Store your final result as `sandusky_0816`.

```{r crop_mask, echo=TRUE, message=TRUE}
aug_16_rast <- terra::rast("erie_raster_ex/images/ts_2017.0816_0822.L4.LE3.CIcyano.MAXIMUM_7day.tif")

#apply first function
clean_aug_rast <- clean_cicyano(aug_16_rast)
#check
terra::freq(clean_aug_rast) #it has the correct interval

#transform the raster
aug_rast_ci <- dn_to_ci(clean_aug_rast)

#crop the raster to sand bb
crop_rast <- terra::crop(aug_rast_ci, erie_sand) 
  
#mask the raster
sandusky_0816 <- terra::mask(crop_rast, erie_sand) 
plot(sandusky_0816)

#check if its cropped to the actual shapefile or a bb
plot(erie_sand)

```

### 2.3 Compare full vs. masked

Using `terra::global()`, calculate and compare the **mean** CI value for:

- The full Lake Erie cleaned/transformed raster
- The Sandusky Bay-masked raster (`sandusky_0816`)

```{r compare_stats, echo=TRUE, message=TRUE}
#lake erie full raster mean CI value
aug_rast_ci %>% terra::global(., "mean", na.rm=TRUE)

#sandusky bay cleaned raster
sandusky_0816 %>% terra::global(., "mean", na.rm=TRUE)

```
The mean value for all of Lake Erie on Aug 16 is about 0.0006 CI where the mean for the Sandusky Bay on Aug 16 is 
0.0002 CI.



### 2.4 Map it

Create a `tmap` map of `sandusky_0816`. Your map should include:

- A meaningful title
- A legend with appropriate labels
- A scale bar

```{r map_sandusky, echo=TRUE, message=TRUE}

tm_shape(sandusky_0816) +
  tm_raster( col.scale = tm_scale_continuous(values = c("#e5f5e0", "#31a354", "#006d2c")),
             col.legend = tm_legend(title="Chlorophyll Cyanobacteria Index",
                                    position = tm_pos_out("right", "center"),
                                    orientation = "portrait")) +
  tm_title("Chlorophyll Cyanobacteria Bloom in Sandusky Ohio - August 16, 2017",
           position = tm_pos_out("center", "top"),
           size=3) +
  tm_scalebar(position = c("left", "bottom"))

```

### Questions for Task 2:

**Q2.1** In your own words, what is the difference between `terra::crop()` and `terra::mask()`? Why did we use both here instead of just one?
-The crop function simply crops a raster to an area of interest, but it does so in a rectangular form. This means that the area is not perfectly bounded to the same shapefile or area of interest, but to a rectangular area around that area of interest. 
-The mask function takes the raster and crops it to the exact boundary of the shapefile or area of interest, not just to a rectangular area surrounding it. We used both to see how a bounding area near sandusky bay would look compared to the exact shapefile outline of the bay (more precise)

**Q2.2** The mean CI value for Sandusky Bay is likely different from the full lake mean. What could explain this? Does this necessarily mean Sandusky Bay has more or less cyanobacteria than average?
- The mean value for Sandusky Bay doesn't necessarily mean there is less than average compared to all of Lake Erie, it may just be because Lake Erie had many more pixel values to take into account with the mean than Sandusky does because it is a smaller area. *************************************************************************************************************************** finish this

---

## Task 3: Multi-temporal comparison

### 3.1 Load and stack multiple rasters

Load and clean **at least four** of the seven rasters (choose dates that span a meaningful time range). Apply `clean_cicyano()` and `dn_to_ci()` to each. Then combine them into a single multi-layer SpatRaster using `c()`.

*Tip: using `purrr::map()` over your `raster_files` vector, combined with your preprocessing functions, can make this more efficient.*

```{r stack_rasters, echo=TRUE, message=TRUE}
aug_02_rast <- terra::rast("erie_raster_ex/images/ts_2017.0802_0808.L4.LE3.CIcyano.MAXIMUM_7day.tif")
aug_23_rast <- terra::rast("erie_raster_ex/images/ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day.tif")
sept_20_rast <- terra::rast("erie_raster_ex/images/ts_2017.0920_0926.L4.LE3.CIcyano.MAXIMUM_7day.tif")
oct_18_rast <- terra::rast("erie_raster_ex/images/ts_2017.1018_1024.L4.LE3.CIcyano.MAXIMUM_7day.tif")

#create list of rasters
raster_files <- list(aug_02_rast, aug_23_rast, sept_20_rast, oct_18_rast)

library(purrr)
cleaned_raster_files <- raster_files %>% 
  purrr::map(~ clean_cicyano(.x)) %>% #applies first function to each in the list
  purrr::map(~ dn_to_ci(.x)) #applies second function to each in the list

#quick check to see each of these steps worked
summary(raster_files[[4]])
summary(cleaned_raster_files[[4]])

#create single multi-layer raster
multi_rast <- rast(cleaned_raster_files) 


```

### 3.2 Temporal summary

Use `terra::global()` on your multi-layer raster to calculate the **mean** CI value per layer (use `na.rm = TRUE`). This will give you one mean value per date.

Organize the results into a data frame with at minimum two columns: `date` and `mean_ci`. You will need to extract the date from each filename. Use `basename()` and/or `stringr::str_extract()` to parse the date string from the file path — the date range is the part of the filename that looks like `0802_0808`.

```{r temporal_summary, echo=TRUE, message=TRUE}
library(stringr)

#calculate mean CI value per layer
CI_value <- terra::global(multi_rast, "mean", na.rm=TRUE)

#create a list of their filenames
file_names <- names(multi_rast) 

#create this dataframe
CI_df <- data.frame(
  file_name = file_names,
  mean_ci = CI_value$mean)

#clean up date names
CI_df <- CI_df %>%
  mutate(start_month = stringr::str_sub(file_name, 9, 10), #extracting month
         start_day = stringr::str_sub(file_name, 11, 12), #extracting day
         end_month = stringr::str_sub(file_name, 14, 15), #find the end month
         end_day = stringr::str_sub(file_name, 16, 17), # find end day
         date= stringr::str_extract(file_name, "\\d+\\.?\\d*")) #plain date
```

### 3.3 Plot the time series

Using `ggplot2`, create a line or point plot showing mean CI values over time. Label your axes clearly. Which date in your dataset shows the highest mean bloom intensity?
---My image from September 20th has the highest mean bloom intensity

```{r time_series_plot, echo=TRUE, message=TRUE}
library(scales) #show the actual CI values

ggplot(data=CI_df) +
  geom_point(aes( x=date, y=mean_ci), color="blue3") +
  labs(x="Date Image Collection Started",
       y="Mean Chlorophyll Cyanobacteria Index per Image",
       title="Average Chlorophyll Cyanobacteria Index per Day - From August to October 2017") +
  scale_y_continuous(labels = label_number(accuracy = 0.0001)) +
  theme_bw()
```

### 3.4 Binary bloom comparison

Choose any **two dates** from your dataset. For each, create a binary raster where cells with valid CI values are coded as `1` and all other cells (NA) as `0`. You will need to think about how to do this using `terra::classify()` or logical raster operations.

Use these binary rasters to create two new rasters:

- **Persistent bloom:** cells where blooms were present on *both* dates
- **New bloom:** cells where blooms were present on the *later* date but NOT the earlier date

Map both results.

*This requires careful logical thinking before writing code. Plan your approach in pseudocode first.*

```{r binary_comparison, echo=TRUE, message=TRUE}
#selecting Aug 23 and Sept 20
clean_aug_23 <- cleaned_raster_files[[2]]
clean_sept_20 <- cleaned_raster_files[[3]]

#formula check for the smallest value that should be in these rasters
10**(((3.0 / 250.0) * 1) - 4.2) #since we limited the values from 1-249
10**(((3.0 / 250.0) * 249) - 4.2) #checking highest value

#new reclass matrix
binary_raster <- function(x) {
  
  rcl_matrix_new <- matrix(c(6.486344e-05,0.0613762,1),ncol = 3, byrow = TRUE)  
  #expected high and low values
  
  #cleaned  up raster, setting invalid data to NA
  binary_raster <- terra::classify(x, rcl = rcl_matrix_new, others=NA)Bu
  
  return(binary_raster)
}

#running binary raster on both dates
final_aug23 <-binary_raster(clean_aug_23)
final_sept20 <- binary_raster(clean_sept_20)

#plotting to see if this worked
plot(final_aug23)
```

### Questions for Task 3:

**Q3.1** What are the limitations of using **mean CI value** as a measure of bloom severity across the whole lake? What would be a more robust alternative metric?
-Some of the limitations would be that it doesn't distinguish between a high bloom or low bloom severity, it could be a small, very intense bloom that is treated the same as a very large weak bloom. Another issue is that using the entire lake could dilute the mean bloom because blooms typically occur along the coastlines and in small shallow portions of the lake. A better metric could be doing % bloom extent and calculating what percent of the lake had a bloom above certain ranges (low, moderate, high) or look at a range of summary statistics and how it relates to the spatial area.

**Q3.2** Stacking rasters with `c()` in terra creates a multi-layer SpatRaster. What requirement must all rasters meet in order to be stacked this way? (Think: extent, resolution, CRS.)
-These rasters must have the same extent, resolution, be projected into the same CRS and cropped to the same areas. If they aren't the rasters could improperly stack and could extend in different areas and have incorrect analysis.
---

## Task 4: Zonal analysis with areas of interest

### 4.1 — Load all boundary files

Use `list.files()` to get all `.shp` files from the `bounds/` directory. Load each with `sf::read_sf()`, then combine them into a single sf object using `dplyr::bind_rows()`. Make sure each polygon has an identifier — if not, add a column with the filename or AOI name.

```{r load_all_aois, echo=TRUE, message=TRUE}
#listing files
list.files(path="./erie_raster_ex/bounds", pattern="\\.shp$")

bb_e_islands <- sf::read_sf("erie_raster_ex/bounds/bb_erie_islands.shp")
bb_e_large <- sf::read_sf("erie_raster_ex/bounds/bb_erie_large.shp")
bb_e_maumee <- sf::read_sf("erie_raster_ex/bounds/bb_erie_maumee.shp")
bb_e_sandusky <- sf::read_sf("erie_raster_ex/bounds/bb_erie_sandusky.shp")
bb_e_sand_plus <- sf::read_sf("erie_raster_ex/bounds/bb_erie_sandusky_plus.shp")
bb_e_west_sister <- sf::read_sf("erie_raster_ex/bounds/bb_erie_west_sister.shp")

#combine them
combined_sf <- dplyr::bind_rows(bb_e_islands, bb_e_large, bb_e_maumee, bb_e_sandusky,
                                bb_e_sand_plus,bb_e_west_sister) %>%
  mutate(AOI = c("e_islands", "e_large", "e_maumee", "e_sandusky", "e_sand_plus", "e_west_sister"))
```

### 4.2 Extract by area

Using the date you identified in Task 3.3 as having the highest mean bloom intensity, extract CI values for each area of interest using `terra::extract()`. Calculate the **mean** and **maximum** CI per AOI.

*Tip: `terra::extract()` returns a data frame. The `ID` column links rows to the AOI that was used for extraction. You can use `group_by()` and `summarize()` to aggregate.*

```{r zonal_extract, echo=TRUE, message=TRUE}
# your code here
```

### 4.3 Summary table

Present your results as a clean summary table. Which AOI had the highest mean CI value on that date? Which had the highest maximum?

```{r summary_table, echo=TRUE, message=TRUE}
# your code here
```

### 4.4 Map with AOI overlays

Create a `tmap` map showing the CI raster for your chosen date with the AOI boundaries overlaid. Use `tm_borders()` for the boundaries and label them if possible.

```{r map_with_aois, echo=TRUE, message=TRUE}
# your code here
```

### Questions for Task 4:

**Q4.1** You may have gotten `NA` values in your `terra::extract()` results for some AOIs. What could cause this? Name at least two reasons.

**Q4.2** Looking at the spatial distribution of your AOIs on the map, which areas do you expect to be most prone to severe blooms, and why? (Consider geography, hydrology, land use in the Lake Erie watershed.)

---


## Questions

1. Throughout this lab you have been working with **raster** data. In what ways was the analysis workflow different from working with vector data (like you did in Labs 1–3)? What concepts translated directly, and what was fundamentally different?

2. Cyanobacteria blooms are affected by temperature, nutrient runoff, and lake circulation. Based on the temporal patterns you observed in Task 3, does the timing of peak bloom match what you would expect ecologically? Explain your reasoning (I know you're not an expert on blooms - just demonstrate thinking).

3. We used `terra::extract()` in Task 4 to perform a zonal summary. How does this relate conceptually to the **zonal operations** category of map algebra discussed in class?


